---
title: "Markov Stability Analysis"
author:
  - name: Mohammed Saqr
    url: https://saqr.me
    affiliation: University of Eastern Finland
  - name: Sonsoles López-Pernas
    url: https://sonsoles.me
    affiliation: University of Eastern Finland
  - name: Kamila Misiejuk
    url: https://kamilamisiejuk.com
    affiliation: FernUniversität in Hagen, Germany
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Markov Stability Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  out.width = "100%",
  fig.width = 7,
  fig.height = 5,
  dpi       = 96,
  warning   = FALSE,
  message   = FALSE
)
```

```{r}
library(Nestimate)
```

Nestimate provides two functions for Markov-chain stability analysis of
transition networks:

- **`passage_time()`** -- computes the full matrix of mean first passage times
  (MFPT). Entry M[i, j] is the expected number of steps to travel from state
  *i* to state *j* for the first time. The diagonal equals the mean recurrence
  time 1/pi.
- **`markov_stability()`** -- computes per-state stability metrics: persistence,
  stationary probability, return time, sojourn time, and mean accessibility.

Both functions accept a `netobject`, `cograph_network`, `tna` object,
row-stochastic matrix, or a raw wide sequence data frame.

---

## Data

`trajectories` contains 138 learners recorded at 15 time-steps each with three
engagement states: *Active*, *Average*, and *Disengaged*.

```{r}
dim(trajectories)
table(as.vector(trajectories), useNA = "always")
```

---

## Selecting mostly-active learners

We keep learners who were *Active* **more than half the time** (> 7 of 15 steps).

```{r}
sub <- trajectories[rowSums(trajectories == "Active", na.rm = TRUE) > 7, ]
nrow(sub)
```

---

## Sequence plots

`sequence_plot()` visualises action trajectories across sessions. With `type = "heatmap"` (the default), rows represent individual actors and columns represent time steps, producing a dense carpet plot reordered by a dendrogram. With `type = "index"`, rows are separated by thin gaps and can be faceted via `ncol` and `nrow`, or grouped using `group` or a `net_clustering` object. With `type = "distribution"`, state proportions (or counts) are tracked across time steps and rendered as a stacked area or bar chart.

Below we plot the full dataset and a subset restricted to students with 7 or more time points (mostly-active students) using `type = "heatmap"`.

```{r fig.height=7}
sequence_plot(trajectories, type = "heatmap")
```

```{r fig.height=5}
sequence_plot(sub, type = "heatmap")
```

---

## Transition networks

Transition networks summarise how likely each state is to be followed by another, with edge weights representing the probability of moving from one state to the next. Here we build networks for both the full sample and the active subgroup to compare how transition structure differs across groups.

```{r}
net_all <- build_network(trajectories, method = "relative")
net_sub <- build_network(sub, method = "relative")

round(net_all$weights, 3)
round(net_sub$weights, 3)
```

In the mostly-active student group nearly every state transitions predominantly back
to *Active*, and the probability of entering *Disengaged* is almost zero.

---

## Mean First Passage Times

`passage_time()` computes how long it takes, on average, to reach one state from another. Each value in the matrix is the expected number of steps to arrive at a given state for the first time, starting from another. 

```{r}
pt_all <- passage_time(net_all)
pt_sub <- passage_time(net_sub)
```

```{r}
print(pt_all, digits = 2)
```

```{r}
print(pt_sub, digits = 2)
```

**Active to Disengaged** rises from 10.4 to ~40 steps -- mostly-active learners
are nearly four times harder to disengage. The diagonal equals the mean
recurrence time (how many steps between consecutive visits to the same state).

```{r fig.height=4}
plot(pt_all, title = "Full sample (n = 138)")
```

```{r fig.height=4}
plot(pt_sub, title = "Mostly-active learners (n = 42)")
```

In the mostly-active heatmap the *Active* column is uniformly dark (quickly
reachable from any state) and the *Disengaged* column is uniformly pale
(nearly unreachable).

---

## Stationary distribution

```{r echo=FALSE}
data.frame(
  State           = names(pt_all$stationary),
  `Full sample`   = paste0(round(pt_all$stationary * 100, 1), "%"),
  `Mostly active` = paste0(round(pt_sub$stationary * 100, 1), "%"),
  check.names     = FALSE
)
```

In the mostly-active group ~79% of long-run time is spent *Active* (vs 37% in
the full sample).

---

## Markov Stability

`markov_stability()` summarises how stable and accessible each state is within a transition network, computing five per-state metrics: persistence (self-loop probability), stationary distribution, mean recurrence time, sojourn time, and mean accessibility to and from other states.


```{r}
ms_all <- markov_stability(net_all)
ms_sub <- markov_stability(net_sub)

print(ms_all)
print(ms_sub)
```

The `stability` data frame contains:

| Column | Meaning |
|--------|---------|
| `persistence` | P[i,i] -- probability of staying at the next step |
| `stationary_prob` | Long-run proportion of time in this state |
| `return_time` | Expected steps between consecutive visits (1/pi) |
| `sojourn_time` | Expected consecutive steps before leaving (1/(1-P[i,i])) |
| `avg_time_to_others` | Mean MFPT leaving this state to all others |
| `avg_time_from_others` | Mean MFPT from all other states arriving here |

Plotting the stability metrics as bar charts allows for a quick visual comparison between the full sample and the mostly-active subgroup.

```{r fig.height=6}
plot(ms_all, title = "Full sample")
```

```{r fig.height=6}
plot(ms_sub, title = "Mostly-active learners")
```
In the mostly-active subgroup, `Active` dominates: it has the highest persistence (0.82) and sojourn time (5.53 steps), meaning learners who enter this state tend to remain there. In the full sample, `Average` is the most reachable state, with an `avg_time_from_others` of just 3.20 steps, but this rises sharply in the mostly-active subgroup (6.20 steps) as `Active` absorbs most transitions. `Disengaged` is effectively unreachable in the mostly-active subgroup, with an `avg_time_from_others` of 51.64 steps and a stationary probability of just 0.024, compared to 0.185 in the full sample.

---
