---
title: "Transition Network Analysis with Nestimate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transition Network Analysis with Nestimate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

Nestimate provides a unified interface for estimating networks from sequence data. This vignette demonstrates transition network estimation, group comparison, and bootstrap inference using the `human_ai_cat` dataset --- coded interaction sequences from 429 human-AI pair programming sessions.

## Data

The `human_cat` dataset contains 10,796 coded human interactions classified into 9 categories.

```{r data}
library(Nestimate)
head(human_cat[, c("session_id", "category")], 8)
table(human_cat$category)
```

## Building networks

`build_network()` accepts long-format data directly. The `action` parameter specifies the state column, `actor` defines sequences, and `time` provides ordering.

### Transition network (TNA)

Row-normalized transition probabilities --- each row sums to 1.

```{r tna}
net_tna <- build_network(human_cat, method = "tna",
                         action = "category", actor = "session_id",
                         time = "id")
net_tna
```

### Frequency network (FTNA)

Raw transition counts --- preserves how often each transition occurs.

```{r ftna}
net_ftna <- build_network(human_cat, method = "ftna",
                          action = "category", actor = "session_id",
                          time = "id")
net_ftna
```

### Attention network (ATNA)

Decay-weighted transitions --- recent transitions receive more weight. The `lambda` parameter controls decay rate.

```{r atna}
net_atna <- build_network(human_cat, method = "atna",
                          action = "category", actor = "session_id",
                          time = "id", lambda = 0.5)
net_atna
```

## Group comparison

The `group` parameter builds separate networks per group. Here we compare human and AI interaction patterns using the full `human_ai_cat` dataset (19,347 interactions).

```{r groups}
nets <- build_network(human_ai_cat, method = "tna",
                      action = "category", actor = "session_id",
                      time = "id", group = "actor")
nets
```

```{r group-detail}
nets$Human
nets$AI
```

## Bootstrap inference

Bootstrap resampling assesses which edges are statistically stable. Sequences are resampled with replacement and the network is re-estimated each iteration.

```{r bootstrap}
boot <- bootstrap_network(nets, iter = 1000, seed = 42)
```

```{r boot-human}
summary(boot$Human)
```

```{r boot-ai}
summary(boot$AI)
```

The pruned model retains only significant edges:

```{r pruned}
boot$Human$model
boot$AI$model
```
