---
title: "Analysis: engagement trajectories and dropout risk"
subtitle: "What chain-structure adds to a 3-state engagement model"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
    df_print: kable
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "",
  out.width  = "100%",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  warning    = FALSE,
  message    = FALSE
)
library(Nestimate)
kt <- function(df, ...) knitr::kable(df, row.names = FALSE, ...)
```

The bundled `trajectories` dataset records the weekly engagement state
of 138 students across 15 weeks of a course. Three states: **Active**,
**Average**, **Disengaged**. Some students have a *terminal-NA*
pattern in the matrix — once their data goes missing, it stays missing
— which we treat as **dropout**. This is the cleanest dataset in the
package for `chain_structure()` because it admits two complementary
chain models:

1. **Engagement-only chain** (3 states): how do students cycle
   between engagement levels? Regular, non-reversible —
   `markov_stability()`'s long-run distribution is interpretable.
2. **Augmented chain with Dropout as a fourth absorbing state**: how
   long until a student drops out, conditional on current
   engagement? The three engagement states are now **transient** and
   `markov_stability()` is *not* the right summary —
   `chain_structure()` is.

Both models are needed for a complete reading. Two functions in the
package handle the wide-data plumbing:

- `actor_endpoints(data)` — tidy per-actor first/last state +
  dropout flag.
- `mark_terminal_state(data, state = "Dropout")` — fill terminal NAs
  with an explicit label so the resulting chain has the label as a
  structurally absorbing state.

---

## Setup

```{r}
ep <- actor_endpoints(trajectories)
kt(data.frame(
  metric = c("students", "weeks observed", "dropouts (terminal NA)",
             "completers"),
  value  = c(nrow(ep), ncol(trajectories),
             sum(ep$dropped_out), sum(!ep$dropped_out)),
  stringsAsFactors = FALSE
))
```

```{r}
kt(head(ep, 6),
   caption = "actor_endpoints() — first six rows")
```

Each row records when the student entered the course (`first_step`,
`first_state`), when they were last seen (`last_step`,
`last_state`), and whether they dropped out (`last_step < 15` with
all subsequent weeks NA).

---

## Part 1 — Engagement chain (no dropout)

`build_network()` on the raw matrix gives the engagement-only chain:
the dynamics conditional on remaining in the course. NAs are skipped
silently when transitions are counted.

```{r}
net_eng <- build_network(as.data.frame(trajectories), method = "relative")
cs_eng  <- chain_structure(net_eng)
summary(cs_eng)
```

Reading the summary: the chain is **regular** (irreducible plus
aperiodic), so `markov_stability()`'s stationary distribution is
interpretable. **Disengaged is the least sticky state**
(persistence ≈ 0.48, sojourn ≈ 1.93 weeks); Active is the stickiest
(persistence ≈ 0.70, sojourn ≈ 3.31). All three states return to
themselves with probability 1 (`return_probability` = 1 for recurrent
states by definition).

Long-run mass: **~37% Active, ~44% Average, ~19% Disengaged**.
Disengaged absorbs the smallest share of student-weeks.

```{r}
kt(round(net_eng$weights, 3),
   caption = "Engagement transition matrix (rows = this week, cols = next)")
```

The off-diagonals show how each state is exited. From Disengaged,
**40% recover to Average and 12% jump to Active**. From Active, only
**3.5% drop directly to Disengaged** — the Active → Disengaged
direct path is ten times less likely than Average → Disengaged,
suggesting disengagement is reached *gradually*, not abruptly.

```{r, fig.height = 4}
plot(cs_eng)
```

The heatmap is saturated dark across all 9 cells: every state is
reachable from every other with probability 1. The diagonal pattern
is the visual signature of non-reversibility — the asymmetric flow
between Active, Average, and Disengaged.

---

## Part 2 — Augmented chain with Dropout

`mark_terminal_state(trajectories, state = "Dropout")` fills every
terminal NA with the label `"Dropout"`. The resulting chain has
Dropout as a structurally absorbing state.

```{r}
M_aug   <- mark_terminal_state(trajectories, state = "Dropout")
net_aug <- build_network(M_aug, method = "relative")
cs_aug  <- chain_structure(net_aug)
summary(cs_aug)
```

The summary now tells a different story:

- **`is_regular = FALSE`** — the chain is not regular because Dropout
  is absorbing. Running `markov_stability()` here would put all
  long-run mass on Dropout (correct as a limit, useless as a
  characterisation of student behaviour).
- **The three engagement states are classified `transient`**, and
  Dropout is `absorbing` with `period = 1` and infinite sojourn
  (once entered, never left).
- **`absorption_probability` is 1 for every transient state** — every
  student eventually drops out under this model. That follows from
  Dropout being the only absorbing state and every transient state
  having a positive path to it.
- **`mean_absorption_time` is what differs by starting state**.

```{r}
kt(round(net_aug$weights, 3),
   caption = "Augmented transition matrix (Dropout column = per-week dropout probability)")
```

The Dropout column reads off the **per-week dropout probability** by
current engagement state:

- **Active → Dropout: ~0.3% per week** (~1 in 333 student-weeks)
- **Average → Dropout: ~0.8% per week**
- **Disengaged → Dropout: ~3.0% per week** (~1 in 33 student-weeks)

Disengaged students drop out at **~10× the rate** of Active students.
A single week of disengagement carries the immediate dropout risk of
ten weeks of active engagement.

```{r, fig.height = 4}
plot(cs_aug)
```

Visually, the **Dropout column is solid dark** (every state reaches
Dropout with probability 1) and the **Dropout row is white except
in the Dropout column** (once dropped, no return). The asymmetry of
that row vs column is the signature of an absorbing state.

### Why does starting state matter so little for absorption time?

The chain is *forgiving*. Even though Disengaged → Dropout per-week
rate is 10× Active → Dropout, students mix between engagement states
fast: from Disengaged, **52% are at Average or Active next week**.
The bad-week-now disadvantage is diluted within one or two steps.

Mathematically, `mean_absorption_time` integrates over future
trajectories, weighted by transition probabilities. A Disengaged
student's expected dropout time is **not** the inverse of the
immediate dropout rate (1/0.030 ≈ 33 weeks); it is the
equilibrium-mixture-weighted rate (~1/0.0104 ≈ 96 weeks). The chain
absorbs the short-run rate ratio into a long-run rate that is much
closer across starting states.

This is a general property of regular transient cores embedded in
absorbing chains — the long-run absorption rate is closer to the
stationary mixture rate than to any individual state's per-step
rate. `chain_structure()` makes both the immediate rates (visible
in the augmented transition matrix) and the absorbed-by rates
(in `mean_absorption_time`) visible side by side.

---

## Part 3 — Cross-tabulation: where were dropouts last seen?

The model-based view above has a model-free counterpart: simply
cross-tabulate the per-actor `last_state` against the dropout flag.

```{r}
tab <- with(ep[!is.na(ep$last_state), ],
            table(last_state, dropped_out, dnn = c("last_state", "dropped_out")))
kt(data.frame(
  last_state  = rownames(tab),
  dropouts    = as.integer(tab[, "TRUE"]),
  completers  = as.integer(tab[, "FALSE"]),
  pct_dropout = sprintf("%.1f%%",
                        100 * tab[, "TRUE"] /
                              (tab[, "TRUE"] + tab[, "FALSE"])),
  stringsAsFactors = FALSE
))
```

**Of students last seen Disengaged, ~35% dropped out**. Of students
last seen Average, ~10% dropped out. Of students last seen Active,
~5% dropped out. The ten-to-one ratio between Disengaged and Active
matches the per-week transition-probability ratio computed via the
chain. The two readings agree on the behavioural signal.

---

## When to use which

| Question | Right summary |
|---|---|
| Long-run engagement balance, ignoring dropout | `markov_stability()` on the engagement-only chain |
| Is the engagement chain regular? Does the stationary distribution exist? | `chain_structure()$is_regular` on the engagement-only chain |
| Per-week dropout probability by current state | `P[i, "Dropout"]` from `build_network(mark_terminal_state(data))` |
| Expected weeks to dropout from each state | `chain_structure()$mean_absorption_time` on the augmented chain |
| Probability of *eventual* dropout | `chain_structure()$absorption_probabilities` (= 1 here) |
| Per-actor first/last state and dropout flag | `actor_endpoints(data)` |
| Backwards check: where were the dropouts last seen? | Cross-tab `last_state` × `dropped_out` from `actor_endpoints()` |

The two chain models — engagement-only and engagement-plus-Dropout —
are complementary. The first describes the dynamics of the
*transient core*; the second embeds those dynamics in a funnel
toward absorption. `chain_structure()` is the diagnostic that
chooses the right summary: when an absorbing state is present,
`markov_stability()` is uninterpretable and the absorption analysis
takes its place.

## References

- Kemeny, J. G. and Snell, J. L. (1976). *Finite Markov Chains*.
  Springer-Verlag.
- Norris, J. R. (1997). *Markov Chains*. Cambridge University Press.
- Spedicato, G. A. (2017). Discrete Time Markov Chains with R.
  *R Journal* 9(2), 84–104.
