---
title: "MCML within-cluster equivalence: build_mcml vs NA-mask + tna::tna"
output: html_document
---

This document verifies, cluster by cluster, that
`build_mcml(..., type = "tna")$clusters[[k]]$weights` is bit-identical to the
transition matrix obtained by

1. masking states from other clusters to `NA` in the original wide sequences,
2. calling `tna::tna()` on the masked sequences.

It also checks that `as_tna(MCMLL)` carries the same row-stochastic weights
into the wrapped netobjects, and that the initial probabilities match
`tna::tna()$inits`.

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(Nestimate)
library(tna)

# Maximum absolute element-wise difference, NA-safe.
matrix_delta <- function(a, b) {
  common <- intersect(rownames(a), rownames(b))
  max(abs(a[common, common] - b[common, common]))
}
```

## 1. Load the data

```{r load}
Interactions <- rio::import("/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/complicated_nw2.csv")
Nodesin      <- rio::import("/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/Codemapping.xlsx")

clusters_list <- split(Nodesin$name, Nodesin$groups)
str(clusters_list)
```

## 2. Build MCML and the wide sequences in one place

`build_mcml()` accepts long-format event-log data directly via the
`actor`/`time`/`action` arguments — it pipes the data through `prepare()`
internally. We also call `prepare()` ourselves so we have the same wide
sequence frame to mask later.

```{r mcml-build}
MCMLL <- build_mcml(
  Interactions,
  clusters = clusters_list,
  actor    = "course_group",
  time     = "timestamp",
  action   = "From",
  type     = "tna"
)

prepared <- prepare(
  Interactions,
  actor  = "course_group",
  time   = "timestamp",
  action = "From"
)
sequences <- prepared$sequence_data
```

`MCMLL$macro$weights` is the cluster-level transition matrix. Each
`MCMLL$clusters[[k]]$weights` is the within-cluster row-stochastic chain we
want to verify.

```{r macro-summary}
round(MCMLL$macro$weights, 3)
```

## 3. Tidy NA-masking helper

For each cluster, replace any state outside the cluster with `NA` across all
time-step columns. `dplyr::across()` does the work in one expression.

```{r mask-fn}
mask_to_cluster <- function(seqs, cluster_nodes) {
  seqs |>
    mutate(across(everything(),
                  ~ if_else(. %in% cluster_nodes, ., NA_character_)))
}
```

## 4. Cluster-by-cluster verification

Each section runs the same three-step recipe:

1. Mask the wide sequences to the focal cluster.
2. Call `tna::tna()` on the masked sequences.
3. Compare the resulting transition matrix and inits to `MCMLL`'s output.

### 4.1 Cognitive

```{r cluster-cognitive}
cl <- "Cognitive"; nodes <- clusters_list[[cl]]; nodes

masked_cog <- mask_to_cluster(sequences, nodes)
tna_cog    <- tna::tna(masked_cog)

# MCML vs tna::tna - transition matrix
mcml_W <- MCMLL$clusters[[cl]]$weights
round(mcml_W, 3)
round(tna_cog$weights, 3)
matrix_delta(mcml_W, tna_cog$weights)

# MCML vs tna::tna - inits
mcml_I <- as_tna(MCMLL)[[cl]]$inits
round(mcml_I, 3); round(tna_cog$inits, 3)
max(abs(mcml_I[names(tna_cog$inits)] - tna_cog$inits))
```

### 4.2 Procedural

```{r cluster-procedural}
cl <- "Procedural"; nodes <- clusters_list[[cl]]; nodes

masked_pro <- mask_to_cluster(sequences, nodes)
tna_pro    <- tna::tna(masked_pro)

mcml_W <- MCMLL$clusters[[cl]]$weights
round(mcml_W, 3); round(tna_pro$weights, 3)
matrix_delta(mcml_W, tna_pro$weights)

mcml_I <- as_tna(MCMLL)[[cl]]$inits
round(mcml_I, 3); round(tna_pro$inits, 3)
max(abs(mcml_I[names(tna_pro$inits)] - tna_pro$inits))
```

### 4.3 Question

```{r cluster-question}
cl <- "Question"; nodes <- clusters_list[[cl]]; nodes

masked_q <- mask_to_cluster(sequences, nodes)
tna_q    <- tna::tna(masked_q)

mcml_W <- MCMLL$clusters[[cl]]$weights
round(mcml_W, 3); round(tna_q$weights, 3)
matrix_delta(mcml_W, tna_q$weights)

mcml_I <- as_tna(MCMLL)[[cl]]$inits
round(mcml_I, 3); round(tna_q$inits, 3)
max(abs(mcml_I[names(tna_q$inits)] - tna_q$inits))
```

### 4.4 Reasoning

```{r cluster-reasoning}
cl <- "Reasoning"; nodes <- clusters_list[[cl]]; nodes

masked_r <- mask_to_cluster(sequences, nodes)
tna_r    <- tna::tna(masked_r)

mcml_W <- MCMLL$clusters[[cl]]$weights
round(mcml_W, 3); round(tna_r$weights, 3)
matrix_delta(mcml_W, tna_r$weights)

mcml_I <- as_tna(MCMLL)[[cl]]$inits
round(mcml_I, 3); round(tna_r$inits, 3)
max(abs(mcml_I[names(tna_r$inits)] - tna_r$inits))
```

### 4.5 Social

```{r cluster-social}
cl <- "Social"; nodes <- clusters_list[[cl]]; nodes

masked_s <- mask_to_cluster(sequences, nodes)
tna_s    <- tna::tna(masked_s)

mcml_W <- MCMLL$clusters[[cl]]$weights
round(mcml_W, 3); round(tna_s$weights, 3)
matrix_delta(mcml_W, tna_s$weights)

mcml_I <- as_tna(MCMLL)[[cl]]$inits
round(mcml_I, 3); round(tna_s$inits, 3)
max(abs(mcml_I[names(tna_s$inits)] - tna_s$inits))
```

### 4.6 Tutor

```{r cluster-tutor}
cl <- "Tutor"; nodes <- clusters_list[[cl]]; nodes

masked_t <- mask_to_cluster(sequences, nodes)
tna_t    <- tna::tna(masked_t)

mcml_W <- MCMLL$clusters[[cl]]$weights
round(mcml_W, 3); round(tna_t$weights, 3)
matrix_delta(mcml_W, tna_t$weights)

mcml_I <- as_tna(MCMLL)[[cl]]$inits
round(mcml_I, 3); round(tna_t$inits, 3)
max(abs(mcml_I[names(tna_t$inits)] - tna_t$inits))
```

## 5. All-clusters summary

```{r summary, message = FALSE}
all_results <- imap_dfr(clusters_list, function(nodes, cl_name) {
  masked  <- mask_to_cluster(sequences, nodes)
  tna_net <- tna::tna(masked)
  mcml_W  <- MCMLL$clusters[[cl_name]]$weights
  mcml_I  <- as_tna(MCMLL)[[cl_name]]$inits
  tibble(
    cluster      = cl_name,
    n_states     = length(nodes),
    delta_weights = matrix_delta(mcml_W, tna_net$weights),
    delta_inits  = max(abs(mcml_I[names(tna_net$inits)] - tna_net$inits))
  )
})
all_results
```

A `delta_weights` and `delta_inits` of `0` for every row confirms each
within-cluster network is exactly the conditional Markov chain
`P(next = j | current = i, both i, j ∈ cluster k)` and that the inits match
`tna::tna()`'s column-1 first-state distribution.
