---
title: "MCML analysis of coded discussion interactions"
output: html_document
---

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(Nestimate)
library(tna)
# cograph is referenced via `cograph::splot()` only, not attached, because
# cograph also exports `build_mcml`, `as_tna`, `cluster_summary`, and
# `centrality` and would mask the Nestimate versions used in this analysis.
```

This document walks through a Multi-Cluster Multi-Level (MCML) analysis on a
coded discussion-forum dataset. Each event is a coded utterance (`From` and
`To` columns). The 26 codes are grouped into six theoretical clusters
(Cognitive, Procedural, Question, Reasoning, Social, Tutor) by an external
code-mapping table. The pipeline is:

1. Load the raw event log and the code-mapping table.
2. Build the MCML in one call, passing the actor / time / action arguments
   for session detection and the `labels` argument so within-cluster nodes
   carry human-readable labels (e.g. `Argument` instead of `c_arg`).
3. Convert to a `netobject_group` via `as_tna()` for downstream analysis.
4. Bootstrap the macro (cluster-level) network for edge stability.
5. Bootstrap one within-cluster network in detail.
6. Compute centralities at both levels.
7. Visualise with `cograph::splot()`.

# 1. Load the data

The event log has one row per coded message; `From` is the speaker's coded
utterance and `course_group` identifies the actor (a course-group thread).
The code-mapping table links each code (`name`) to a cluster (`groups`) and
to a human-readable label (`Code`).

```{r load}
Interactions <- rio::import(
  "/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/complicated_nw2.csv"
)
Nodesin <- rio::import(
  "/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/Codemapping.xlsx"
)

clusters_list <- split(Nodesin$name, Nodesin$groups)
labels_df     <- Nodesin |> select(name, Code)

length(clusters_list)         # number of clusters
sapply(clusters_list, length) # nodes per cluster
```

# 2. Build the MCML

`build_mcml()` accepts long-format event-log data directly and pipes it
through `prepare()` for session detection. We pass the `labels` argument so
within-cluster nodes carry the human-readable labels from `Codemapping`.
`type = "tna"` requests row-stochastic transition probabilities at every
level (the within-cluster networks are then conditional Markov chains
`P(next = j | current = i, both i, j ∈ cluster k)`).

```{r build}
MCMLL <- build_mcml(
  Interactions,
  clusters = clusters_list,
  actor    = "course_group",
  time     = "timestamp",
  action   = "From",
  type     = "tna",
  labels   = labels_df
)

MCMLL
```

# 3. The macro (cluster-level) network

The macro network reads `P(next ∈ cluster_j | current ∈ cluster_i)`. The
diagonal is the probability of staying in the same cluster on the next
move; the off-diagonal is the probability of crossing between clusters.

```{r macro-weights}
round(MCMLL$macro$weights, 3)
round(MCMLL$macro$inits,   3)
```

The off-diagonal mass tells you which cluster pairs interact most. The
`inits` vector reads as: out of all session-opening codes, which cluster
opened them.

# 4. Within-cluster networks

Each `MCMLL$clusters[[k]]` is a row-stochastic chain on the nodes of
cluster `k`, with the human-readable labels we passed in. We don't loop —
just inspect a couple to get a feel for the data.

```{r within-cognitive}
round(MCMLL$clusters$Cognitive$weights, 3)
round(MCMLL$clusters$Cognitive$inits,   3)
```

```{r within-reasoning}
round(MCMLL$clusters$Reasoning$weights, 3)
round(MCMLL$clusters$Reasoning$inits,   3)
```

# 5. Convert to a `netobject_group`

`as_tna()` wraps every layer (macro plus one per cluster) as a netobject
suitable for the rest of the package's workflow (bootstrap, centrality,
cograph visualisation). Initial probabilities propagate through.

```{r as-tna}
MCMLL_tna <- as_tna(MCMLL)
class(MCMLL_tna)
names(MCMLL_tna)
```

# 6. Bootstrap the macro network

`bootstrap_network()` resamples the recoded sequences and recomputes the
transition matrix on every iteration, returning per-edge means, 95% CIs,
and a CS-style stability flag (Epskamp, Borsboom & Fried, 2018).

```{r boot-macro, fig.width = 7, fig.height = 5}
set.seed(2026)
boot_macro <- bootstrap_network(MCMLL_tna$macro, iter = 500)
boot_macro

cograph::plot_bootstrap_forest(boot_macro)
```

The print summary tells you how many cluster-to-cluster edges survived the
stability criterion at the chosen confidence level. The forest plot shows
each edge's mean weight with the bootstrap CI.

# 7. Bootstrap a within-cluster network

The Cognitive cluster has four sub-codes. Bootstrapping at this level tests
which within-cluster transitions are stable.

```{r boot-cognitive, fig.width = 7, fig.height = 5}
set.seed(2026)
boot_cog <- bootstrap_network(MCMLL_tna$Cognitive, iter = 500)
boot_cog

# A within-cluster bootstrap may have few or no edges surviving the default
# stability criterion. The forest plot only renders if there is at least one
# significant edge, so we guard the call.
if (isTRUE(any(boot_cog$significant))) {
  cograph::plot_bootstrap_forest(boot_cog)
} else {
  message("No within-cluster edges met the default stability criterion ",
          "for the Cognitive cluster — forest plot skipped.")
}
```

# 8. Centralities

`cograph::centrality()` computes the standard set of network centrality
measures (prestige, brokerage, etc.) on a netobject. We run it on the macro
and on each within-cluster netobject.

```{r centrality-macro}
cograph::centrality(MCMLL_tna$macro)
```

```{r centrality-cognitive}
cograph::centrality(MCMLL_tna$Cognitive)
```

# 9. Visualisation

`cograph::splot()` renders any netobject. Use it on the macro for the
cluster-of-clusters view and on a within-cluster netobject for the
fine-grained view.

```{r splot-macro, fig.width = 7, fig.height = 5}
cograph::splot(MCMLL_tna$macro,
               edge_labels = TRUE,
               edge_label_size = 0.7)
```

```{r splot-cognitive, fig.width = 7, fig.height = 5}
cograph::splot(MCMLL_tna$Cognitive,
               edge_labels = TRUE,
               edge_label_size = 0.7)
```

# 10. What to read off this analysis

* **Macro diagonal vs off-diagonal**: how cluster-bound the discussion is.
  A high `Reasoning -> Reasoning` value means reasoning utterances are
  followed by more reasoning, not by social or procedural follow-ups.
* **Macro inits**: which cluster opens sessions. If the discussion is
  Tutor-initiated, expect a high `Tutor` init.
* **Bootstrap CIs**: a wide CI on a macro edge tells you the cluster-level
  flow is sensitive to which actors / sessions you sample. Narrow CIs
  with stars are the stable interactions worth interpreting.
* **Within-cluster networks**: the conditional dynamics inside each
  cluster, useful for fine-grained pedagogical claims (e.g.
  `Argument -> Counter-argument` rates inside Cognitive).
* **Centralities**: at the macro level, an outgoing-strength hub is a
  cluster the discussion frequently leaves; an incoming hub is a cluster
  it frequently lands in.
