---
title: "MCML within-cluster equivalence: build_mcml vs NA-mask + tna::tna"
output: html_document
---

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(Nestimate)
library(tna)
```

Each within-cluster network produced by `build_mcml(..., type = "tna")` is the
conditional Markov chain on cluster `k`,
`P(next = j | current = i, both i, j ∈ cluster k)`. Equivalent to:

1. take the wide sequences,
2. replace every state outside the cluster with `NA`,
3. call `tna::tna()` on the masked sequences.

This document verifies both the transition matrix and the initial
probabilities for every cluster, against `tna::tna()` directly. Both objects
must be reordered onto the same node order before comparison: `build_mcml()`
preserves the user-supplied cluster order, while `tna::tna()` sorts states
alphabetically.

# 1. Load data

```{r load}
Interactions <- rio::import("/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/complicated_nw2.csv")
Nodesin      <- rio::import("/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/Codemapping.xlsx")

clusters_list <- split(Nodesin$name, Nodesin$groups)
clusters_list
```

# 2. Build MCML and the wide sequences

```{r build}
MCMLL <- build_mcml(
  Interactions,
  clusters = clusters_list,
  actor    = "course_group",
  time     = "timestamp",
  action   = "From",
  type     = "tna"
)

sequences <- prepare(
  Interactions,
  actor  = "course_group",
  time   = "timestamp",
  action = "From"
)$sequence_data

MCMLL_tna <- as_tna(MCMLL)
```

```{r macro-weights}
round(MCMLL$macro$weights, 3)
round(MCMLL$macro$inits,   3)
```

# 3. Cluster-by-cluster verification

Same recipe in every section:

* mask the wide sequences to the focal cluster with
  `mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))`,
* call `tna::tna()`,
* reorder `tna::tna()`'s output onto the cluster's node order,
* compare to MCML's matrix and inits.

## 3.1 Cognitive

```{r cognitive}
nodes <- clusters_list$Cognitive

masked  <- sequences |>
  mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))
tna_net <- tna::tna(masked)

tna_W <- tna_net$weights[nodes, nodes]
tna_I <- tna_net$inits[nodes]

round(MCMLL$clusters$Cognitive$weights, 3)
round(tna_W, 3)
all.equal(MCMLL$clusters$Cognitive$weights, tna_W)

round(MCMLL_tna$Cognitive$inits, 3)
round(tna_I, 3)
all.equal(MCMLL_tna$Cognitive$inits, tna_I)
```

## 3.2 Procedural

```{r procedural}
nodes <- clusters_list$Procedural

masked  <- sequences |>
  mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))
tna_net <- tna::tna(masked)

tna_W <- tna_net$weights[nodes, nodes]
tna_I <- tna_net$inits[nodes]

round(MCMLL$clusters$Procedural$weights, 3)
round(tna_W, 3)
all.equal(MCMLL$clusters$Procedural$weights, tna_W)

round(MCMLL_tna$Procedural$inits, 3)
round(tna_I, 3)
all.equal(MCMLL_tna$Procedural$inits, tna_I)
```

## 3.3 Question

```{r question}
nodes <- clusters_list$Question

masked  <- sequences |>
  mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))
tna_net <- tna::tna(masked)

tna_W <- tna_net$weights[nodes, nodes]
tna_I <- tna_net$inits[nodes]

round(MCMLL$clusters$Question$weights, 3)
round(tna_W, 3)
all.equal(MCMLL$clusters$Question$weights, tna_W)

round(MCMLL_tna$Question$inits, 3)
round(tna_I, 3)
all.equal(MCMLL_tna$Question$inits, tna_I)
```

## 3.4 Reasoning

```{r reasoning}
nodes <- clusters_list$Reasoning

masked  <- sequences |>
  mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))
tna_net <- tna::tna(masked)

tna_W <- tna_net$weights[nodes, nodes]
tna_I <- tna_net$inits[nodes]

round(MCMLL$clusters$Reasoning$weights, 3)
round(tna_W, 3)
all.equal(MCMLL$clusters$Reasoning$weights, tna_W)

round(MCMLL_tna$Reasoning$inits, 3)
round(tna_I, 3)
all.equal(MCMLL_tna$Reasoning$inits, tna_I)
```

## 3.5 Social

```{r social}
nodes <- clusters_list$Social

masked  <- sequences |>
  mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))
tna_net <- tna::tna(masked)

tna_W <- tna_net$weights[nodes, nodes]
tna_I <- tna_net$inits[nodes]

round(MCMLL$clusters$Social$weights, 3)
round(tna_W, 3)
all.equal(MCMLL$clusters$Social$weights, tna_W)

round(MCMLL_tna$Social$inits, 3)
round(tna_I, 3)
all.equal(MCMLL_tna$Social$inits, tna_I)
```

## 3.6 Tutor

```{r tutor}
nodes <- clusters_list$Tutor

masked  <- sequences |>
  mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))
tna_net <- tna::tna(masked)

tna_W <- tna_net$weights[nodes, nodes]
tna_I <- tna_net$inits[nodes]

round(MCMLL$clusters$Tutor$weights, 3)
round(tna_W, 3)
all.equal(MCMLL$clusters$Tutor$weights, tna_W)

round(MCMLL_tna$Tutor$inits, 3)
round(tna_I, 3)
all.equal(MCMLL_tna$Tutor$inits, tna_I)
```

# 4. All-clusters summary

```{r summary, message = FALSE}
imap_dfr(clusters_list, function(nodes, cl) {
  masked  <- sequences |>
    mutate(across(everything(), ~ if_else(. %in% nodes, ., NA_character_)))
  tna_net <- tna::tna(masked)
  tna_W   <- tna_net$weights[nodes, nodes]
  tna_I   <- tna_net$inits[nodes]

  tibble(
    cluster       = cl,
    n_states      = length(nodes),
    delta_weights = max(abs(MCMLL$clusters[[cl]]$weights - tna_W)),
    delta_inits   = max(abs(MCMLL_tna[[cl]]$inits - tna_I))
  )
})
```

A `delta_weights` and `delta_inits` of `0` for every row confirms each
within-cluster network is exactly the conditional Markov chain
`P(next = j | current = i, both i, j ∈ cluster k)`, with initial
probabilities matching `tna::tna()`'s column-1 first-state distribution.
