---
title: "Transition entropy of AI assistant interaction sequences"
author: "Nestimate"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 8,
  fig.height = 4.5,
  dpi      = 96
)
suppressPackageStartupMessages({
  library(Nestimate)
  library(ggplot2)
})
```

## The analytic question

Given the AI assistant has just produced a coded action — *Plan*, *Execute*,
*Repair*, ... — how predictable is what comes next? Two summaries answer
this at different scales. Per-state row entropy $H(P_{i\cdot})$ measures
how dispersed the outgoing transitions from state $i$ are. The chain-level
**entropy rate** $h(P) = \sum_i \pi_i H(P_{i\cdot})$ aggregates row
entropies under the long-run stationary distribution $\pi$ and is the
Shannon–McMillan–Breiman per-step uncertainty of trajectories
[Cover & Thomas, 2006, ch. 4]. The deficit $H(\pi) - h(P)$ quantifies how
much the order-1 Markov structure compresses the process below the
i.i.d. baseline.

## Data

`ai_long` contains 8,551 coded AI-assistant turns across 428 sessions and
34 projects, classified into 8 fine-grained codes and 3 super-class
clusters (Action / Communication / Repair).

```{r load-data}
data(ai_long, package = "Nestimate")
str(ai_long)
table(code = ai_long$code)
```

## Building the transition network

`build_network()` consumes the long-format data directly, identifies the
session boundary via `id_col`, and orders events by `time`. The
`"relative"` method row-normalises raw transition counts to produce a
proper Markov transition matrix.

```{r build-net}
net_codes <- build_network(
  ai_long,
  method  = "relative",
  id_col  = "session_id",
  action  = "code",
  time    = "order_in_session"
)
round(net_codes$weights, 3)
```

Each row sums to one, so `net_codes$weights` is the maximum-likelihood
estimator of the AI's order-1 transition matrix.

## Per-state branching: which codes are most predictable?

```{r te-codes}
te_codes <- transition_entropy(net_codes, base = 2)
te_codes
```

Per-state row entropy is bounded above by $\log_2 8 = 3$ bits (uniform
branching across all 8 codes) and below by 0 (deterministic next code).

```{r summary-codes}
summary(te_codes)
```

The summary table sorts by stationary probability when read alongside
`stationary` — *Execute* is the most-visited state (44 % of stationary
mass), and its row entropy of `r round(te_codes$row_entropy["Execute"], 2)`
bits sits below the chain mean, meaning the AI's behaviour after an
*Execute* turn is more constrained than after, e.g., *Explain*.

```{r plot-codes, fig.height = 4.5}
plot(te_codes, title = "AI codes — per-state branching entropy")
```

Bar widths in the plot are proportional to stationary probability, so the
total visual area sums to the entropy rate $h(P)$. The dashed orange line
is $h(P)$, the dotted grey line is the uniform-branching upper bound
$\log_2 n$.

## Chain-level summary

```{r chain-summary}
te_codes$entropy_rate
te_codes$stationary_entropy
te_codes$redundancy
te_codes$redundancy / te_codes$stationary_entropy
```

Entropy rate $h(P) =$ `r round(te_codes$entropy_rate, 3)` bits/step;
stationary entropy $H(\pi) =$ `r round(te_codes$stationary_entropy, 3)`
bits. The redundancy ratio
$\big(H(\pi) - h(P)\big) / H(\pi) =$
`r round(te_codes$redundancy / te_codes$stationary_entropy, 3)` says the
order-1 Markov memory removes about 8 % of the uncertainty an i.i.d.
process at the same marginals would carry. The chain has structure, but
a single-step memory is far from a deterministic predictor.

The maximum theoretical entropy rate at this state-space cardinality is
$\log_2 8 = 3$ bits. The observed `r round(te_codes$entropy_rate, 2)`
bits represents
`r round(100 * te_codes$entropy_rate / log2(8), 1)` % of that maximum —
the AI's coded action stream is closer to "moderately uncertain" than to
"random".

## Coarsening to 3 super-class clusters

When the same sequences are re-coded at the cluster level (Action,
Communication, Repair), the state space shrinks from 8 to 3 and so does
the maximum entropy ($\log_2 3 \approx 1.585$).

```{r te-clusters}
net_clusters <- build_network(
  ai_long,
  method  = "relative",
  id_col  = "session_id",
  action  = "cluster",
  time    = "order_in_session"
)
te_clusters <- transition_entropy(net_clusters, base = 2)
te_clusters

c(
  fine   = te_codes$entropy_rate    / log2(8),
  coarse = te_clusters$entropy_rate / log2(3)
)
```

Normalising each entropy rate by the appropriate $\log_2 n$ ceiling
removes the trivial state-space-size effect. The two ratios show how
much of the maximum uncertainty each granularity actually attains. A
sharp drop on coarsening would indicate the cluster scheme is capturing
real grouping structure; a flat ratio indicates the cluster boundaries
are essentially arbitrary cuts through a homogeneous transition surface.

```{r plot-clusters, fig.height = 3.5}
plot(te_clusters, title = "AI clusters — per-state branching entropy")
```

## Group dispatch: per-project entropy rates

`transition_entropy()` dispatches on `netobject_group` so a per-project
analysis is one extra argument to `build_network()`.

```{r te-projects, fig.height = 4.5}
projects_keep <- names(which(table(ai_long$project) >= 100))
ai_subset <- ai_long[ai_long$project %in% projects_keep, ]

net_by_project <- build_network(
  ai_subset,
  method = "relative",
  id_col = "session_id",
  action = "code",
  time   = "order_in_session",
  group  = "project"
)
te_by_project <- transition_entropy(net_by_project, base = 2)
te_by_project
```

```{r plot-projects, fig.height = 4}
rates <- vapply(te_by_project, function(o) o$entropy_rate, numeric(1))
df_rates <- data.frame(
  project      = names(rates),
  entropy_rate = unname(rates)
)
df_rates <- df_rates[order(df_rates$entropy_rate), ]
df_rates$project <- factor(df_rates$project, levels = df_rates$project)

ggplot(df_rates, aes(x = entropy_rate, y = project)) +
  geom_col(fill = "#0072B2", alpha = 0.85) +
  geom_vline(xintercept = te_codes$entropy_rate,
             linetype = "dashed", colour = "#D55E00") +
  annotate("text", x = te_codes$entropy_rate, y = 1,
           label = sprintf("Pooled h(P) = %.2f", te_codes$entropy_rate),
           hjust = -0.05, vjust = -0.4, colour = "#D55E00", size = 3.4) +
  labs(x = "Entropy rate (bits)", y = NULL,
       title = "Per-project entropy rate of AI code transitions",
       subtitle = sprintf("%d projects with >=100 turns; vertical line = pooled chain",
                         nrow(df_rates))) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

The spread across projects measures how strongly project context shapes
the AI's order-1 behaviour. Projects clustered tightly around the pooled
$h(P)$ would indicate a stable, project-invariant behavioural profile;
wide spread indicates project-specific transition regimes that a single
pooled model would average over.

## Validation

`transition_entropy()` is validated to machine precision against an
independent reference composition built from
`markovchain::steadyStates()` and `entropy::entropy.plugin()`. Across
100 random irreducible aperiodic configurations spanning 3–7 states with
log bases 2, $e$, 10:

| Quantity              | Max abs error | Tolerance |
|-----------------------|---------------|-----------|
| Entropy rate $h(P)$   | 6.66 × 10⁻¹⁶ | 1 × 10⁻¹⁰ |
| Stationary $H(\pi)$   | 8.88 × 10⁻¹⁶ | 1 × 10⁻¹⁰ |
| Per-row $H(P_{i\cdot})$ | 8.88 × 10⁻¹⁶ | 1 × 10⁻¹⁰ |
| Stationary $\pi$      | 4.44 × 10⁻¹⁶ | 1 × 10⁻⁸  |

Reproduce with
`NESTIMATE_EQUIV_TESTS=true Rscript -e 'devtools::load_all("."); testthat::test_file("local_testing_and_equivalence/test-equiv-transition_entropy.R")'`.

## When to use which

| Quantity              | Reads as                                                               |
|-----------------------|------------------------------------------------------------------------|
| `row_entropy[i]`      | uncertainty of the AI's *next* code given it just emitted code $i$     |
| `stationary_entropy`  | i.i.d. baseline; the entropy of the marginal-state distribution        |
| `entropy_rate`        | irreducible per-step trajectory uncertainty under the order-1 chain    |
| `redundancy`          | bits of memory the chain captures at order 1; zero iff i.i.d.          |

Pair `transition_entropy()` with `markov_order_test()` to ask whether the
order-1 entropy rate is the *true* asymptotic rate or whether longer
memory still removes uncertainty. Pair it with `chain_structure()` to
verify the chain is regular before interpreting $\pi$ as a stable long-run
distribution; for non-regular chains $\pi$ is non-unique and the entropy
rate is conditional on the recurrent class assumed.

## References

Cover, T.M., Thomas, J.A. (2006). *Elements of Information Theory*, 2nd
ed., chapter 4. Wiley.

Shannon, C.E. (1948). A mathematical theory of communication. *Bell
System Technical Journal*, **27**, 379–423.
