---
title: "Transition entropy of student engagement trajectories"
author: "Nestimate"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_depth: 2
    toc_float: true
    code_folding: show
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo     = TRUE,
  message  = FALSE,
  warning  = FALSE,
  fig.width  = 8,
  fig.height = 4.5,
  dpi      = 96
)
suppressPackageStartupMessages({
  library(Nestimate)
  library(ggplot2)
})
```

## The analytic question

`trajectories` records weekly engagement state for 138 students over 15
weeks, in three categories — **Active**, **Average**, **Disengaged**.
Some students stop being observed before week 15; the trailing NAs in
the matrix are the structural signature of dropout.

Two questions:

1. How predictable is a student's *next* engagement state given the
   current one? At what fraction of the theoretical maximum does the
   chain operate?
2. When dropout is treated as an explicit absorbing state, what does
   the entropy rate become — and why?

The same $h(P)$ summary answers both questions but produces dramatically
different numbers, and the gap is the substantive point of this
tutorial.

## Data

```{r data}
data(trajectories, package = "Nestimate")
dim(trajectories)
head(trajectories[, 1:8])
table(unlist(trajectories), useNA = "ifany")
```

200 cells are NA — these are the dropout signatures. The
*engagement-only* analysis below treats them as missing; the *augmented*
analysis below replaces them with an explicit `"Dropout"` label.

## Engagement-only chain (3 states)

`build_network(method = "relative")` row-normalises the empirical
transition counts ignoring NA destinations.

```{r net-engage}
net_eng <- build_network(trajectories, method = "relative")
round(net_eng$weights, 3)
```

Two patterns are visible by inspection: the diagonal is the dominant
mass everywhere (engagement is sticky), and the row for Disengaged is
the flattest (most spread of outgoing probability).

```{r te-engage}
te_eng <- transition_entropy(net_eng, base = 2)
te_eng
```

Each numeric quantity is reported in two forms: the raw entropy in
bits, and the normalised entropy obtained by dividing by the ceiling
$\log_2 n$ (here $\log_2 3 \approx 1.585$). The normalised forms live
in $[0, 1]$ so they are directly comparable across chains with
different state counts: 0 means deterministic, 1 means uniform
branching across all states. The raw bits remain the right unit when
the analyst needs an information-theoretic quantity (compression
limits, channel capacity, model-selection penalties).

The chain operates at **`r round(100 * te_eng$entropy_rate_norm, 1)` %**
of its theoretical maximum entropy. The redundancy is **`r round(100 * te_eng$redundancy_norm, 1)` %**
of the marginal uncertainty $H(\pi)$ — the order-1 Markov memory
removes about a sixth of what an i.i.d. process at the same marginals
would carry.

```{r te-engage-plot, fig.height = 4}
plot(te_eng, title = "Engagement chain — per-state branching entropy")
```

### Per-state contribution

The entropy rate decomposes additively. Each state contributes
$\pi_i \cdot H(P_{i\cdot})$ to the chain rate. `summary()` returns a
tidy data.frame sorted by share of $h(P)$, descending:

```{r contrib}
summary(te_eng)
```

Reading the table top-to-bottom:

- **Average dominates the chain's uncertainty budget at ~48 %** of
  $h(P)$, both because it is the most-visited state ($\pi \approx 0.44$)
  *and* because its row entropy is mid-range
  (normalised `r round(te_eng$row_entropy_norm["Average"], 2)`).
- **Active contributes ~31 %** despite being almost as frequent — its
  normalised row entropy is only `r round(te_eng$row_entropy_norm["Active"], 2)`,
  meaning once a student is Active, the next state is comparatively
  predictable.
- **Disengaged has the most spread-out outgoing distribution**
  (normalised row entropy `r round(te_eng$row_entropy_norm["Disengaged"], 2)`,
  near the ceiling), but contributes only ~21 % to the total because
  it is rarely occupied.

This decomposition is what `transition_entropy()` adds over a single
scalar summary: it says *where* the chain spends its uncertainty
budget, not just how much. The normalised row entropy column makes
the within-row comparison scale-free; the contribution percent makes
the across-row comparison scale-free.

## Augmented chain with explicit Dropout (4 states)

`mark_terminal_state()` fills trailing NAs with an explicit `"Dropout"`
label, turning the censoring pattern into a fourth, *structurally
absorbing*, state.

```{r net-aug}
df_aug  <- mark_terminal_state(trajectories,
                                state = "Dropout")
net_aug <- build_network(df_aug, method = "relative")
round(net_aug$weights, 3)
```

The new row for Dropout is `(0, 0, 0, 1)` — a deterministic self-loop.
This is the structural definition of an absorbing state.

```{r te-aug}
te_aug <- transition_entropy(net_aug, base = 2)
te_aug
```

**The entropy rate is zero.** This is a feature of the chain, not a
bug. The stationary distribution of an absorbing chain places all mass
on the absorbing state, so

$$h(P) = \sum_i \pi_i H(P_{i\cdot}) = 1 \cdot H(P_{\text{Dropout}\cdot}) + 0 + 0 + 0 = 0.$$

Every student eventually drops out, after which the chain is
deterministic, so the asymptotic per-step uncertainty is exactly zero.
The per-state row entropies still tell the substantive story:

```{r aug-rows}
data.frame(
  state            = names(te_aug$row_entropy),
  row_entropy      = round(te_aug$row_entropy, 3),
  stationary       = round(te_aug$stationary,  3)
)
```

The transient states (Active, Average, Disengaged) carry exactly the
same row entropies as in the engagement-only chain (modulo the small
probability of dropout), but their stationary mass is zero. The 4-state
$h(P) = 0$ is mathematically correct and substantively the wrong
question — it tells you the long-run fate, not the engagement dynamics.

### Conditional entropy in the transient core

For an absorbing chain, the substantively meaningful quantity is the
entropy rate **conditional on not yet being absorbed** — and that
quantity is what the engagement-only chain already estimates:
$h_{\text{transient}} = `r round(te_eng$entropy_rate, 3)`$ bits/step.
The two views are complementary:

- 4-state augmented $h(P) = 0$ — eventually deterministic
- 3-state engagement-only $h(P) = `r round(te_eng$entropy_rate, 3)`$ —
  per-step uncertainty *while engagement is observed*

`chain_structure()` formalises which view to use:

```{r chain-struct}
cs_eng <- chain_structure(net_eng)
cs_aug <- chain_structure(net_aug)

data.frame(
  chain        = c("engagement-only", "augmented"),
  is_regular   = c(cs_eng$is_regular,   cs_aug$is_regular),
  is_absorbing = c(length(cs_eng$absorbing_states) > 0,
                   length(cs_aug$absorbing_states) > 0),
  h_P          = c(round(te_eng$entropy_rate, 3),
                   round(te_aug$entropy_rate, 3))
)
```

When `is_regular` is `TRUE` the entropy rate is the canonical
information-theoretic summary. When `is_absorbing` is `TRUE` the rate
is mechanical (always zero) and the row entropies of the transient
states are what the analyst should report.

## Is order-1 the right resolution?

The entropy rate is a *first-order* Markov summary. It is the true
asymptotic per-step uncertainty only if the chain is genuinely
order-1. `markov_order_test()` decides.

```{r mo-test}
mo <- markov_order_test(trajectories,
                        max_order = 3, n_perm = 200)
mo
```

The permutation-LRT rejects order 1 *and* order 2 against higher
alternatives ($p < 0.005$ at every order tested), which means the
empirical $h_1 = `r round(te_eng$entropy_rate, 3)`$ bits **overstates**
the true asymptotic entropy rate $h_\infty$. BIC selects order 2,
suggesting the over-estimate at order 1 is small but real. In practice
this means weekly engagement carries memory beyond the immediately
preceding state — an Active week followed by an Average week predicts
the third week differently than two consecutive Average weeks.

When `markov_order_test()` selects an order $k > 1$, the order-1
$h(P)$ should be reported with the caveat that $h_\infty < h_1$, and
the magnitude of the gap should be estimated from $h_2 - h_1$ on the
log-likelihood column of the test table:

```{r h-drop}
ll  <- mo$test_table$loglik
n_obs <- sum(!is.na(trajectories[, -1]))
h1  <- -ll[mo$test_table$order == 1] / n_obs / log(2)  # bits/step
h2  <- -ll[mo$test_table$order == 2] / n_obs / log(2)
data.frame(
  order_1 = round(h1, 3),
  order_2 = round(h2, 3),
  drop    = round(h1 - h2, 3),
  drop_pct = round(100 * (h1 - h2) / h1, 1)
)
```

The order-2 entropy rate is `r round(h2, 3)` bits/step. The
$`r round(100 * (h1 - h2) / h1, 1)`$ % drop from order-1 is the
fraction of "apparent unpredictability" that vanishes once the
preceding pair of states is known — the bits of memory at order 2.

## Where does the order-2 information actually live?

The chain-level $h_1 - h_2$ quantifies *how much* extra information
order-2 carries on average — but does not say *where*. Are all contexts
slightly more predictable, or is the chain genuinely Markov in most
contexts and dramatically non-Markov in a few? `path_dependence()`
decomposes the chain-level drop per context.

For each ordered pair $(s_1, s_2)$ that occurs at least `min_count`
times, it computes the empirical conditional $P_2(\cdot \mid s_1, s_2)$,
compares it against the order-1 prediction $P_1(\cdot \mid s_2)$, and
reports the Kullback-Leibler divergence
$D_{KL}(P_2 \,\Vert\, P_1)$ — the bits of "surprise" you incur if you
use the shorter context when the longer one is the truth. Contexts where
the *modal* (most-likely) next state changes between orders are flagged.

```{r pd-order2}
pd2 <- path_dependence(trajectories,
                       order = 2L, min_count = 5L)
pd2
```

The chain-level **weighted KL = `r round(pd2$chain$KL_weighted, 3)`** bits is
small *on average*, but the per-context table reveals one striking
asymmetry: **`Active -> Disengoged`** has the highest KL among order-2
contexts and *flips the modal next state* — order-1 says the most likely
next state is Disengaged again, but order-2 says it is Active. A
disengagement *coming from* Active is qualitatively different from a
disengagement that has been ongoing: students appear to recover.

```{r pd-order2-plot, fig.height = 4.2}
plot(pd2)
```

The lollipop chart sorts contexts by KL. Orange points are
modal-prediction flips — the substantively interesting contexts where
longer history changes *what you would predict*, not just how confident
you are.

### Order 3: trajectory-level patterns

Going one step further reveals trajectory-level signatures. With
`order = 3` the context becomes a length-2 history $(s_1, s_2)$
predicting $s_3$:

```{r pd-order3}
pd3 <- path_dependence(trajectories,
                       order = 3L, min_count = 5L)
print(pd3, top = 8)
```

Three trajectory patterns emerge from the top contexts:

- **Recovery sequences** — `Disengaged -> Average -> Active`
  (KL `r round(pd3$contexts$KL[pd3$contexts$context == "Disengaged -> Average -> Active"][1], 2)` bits): order-1 predicts continued Active,
  but the full recovery trajectory shows a tendency to *fall back* to
  Average. The relapse risk is invisible to a Markov model.
- **Relapse from Disengaged** — `Active -> Disengaged -> Average`
  (KL `r round(pd3$contexts$KL[pd3$contexts$context == "Active -> Disengaged -> Average"][1], 2)` bits): the path through Disengaged from
  Active predicts Disengaged again, not Average's usual return-to-mean.
- **Confirmed disengagement** — `Disengaged -> Disengaged -> Average`
  (KL `r round(pd3$contexts$KL[pd3$contexts$context == "Disengaged -> Disengaged -> Average"][1], 2)` bits, large H_drop): two consecutive
  Disengaged weeks then Average is much more *predictable* than the
  order-1 baseline suggests, in the direction of "back to Disengaged".

```{r pd-order3-plot, fig.height = 5}
plot(pd3, top = 12, title = "Order-3 trajectory dependence")
```

### Reading the diagnostic

`path_dependence()` answers a different question than
`markov_order_test()`. The order test answers "**should** I use longer
memory globally?" The path-dependence diagnostic answers "**where**
does longer memory actually buy me predictive power?" — and identifies
the substantively meaningful trajectories on which any higher-order
model improves over the chain.

If the goal is prediction, the per-context KL identifies the high-value
contexts where moving to a higher-order model matters most. If the goal
is interpretation, the modal-flip flag identifies the contexts where a
clinician / instructor should treat a learner differently than the
order-1 model would suggest.

## How does this differ from n-grams, MOgen, and HYPA?

`path_dependence()` enters a crowded landscape. Three older tools in
this package — `path_counts()`, `build_mogen()`, `build_hypa()` — also
operate on k-step paths through a categorical sequence. A reader who has
used any of them is right to ask whether the new diagnostic is
genuinely different or just a rebadged version of work the field has
already done. This section locates `path_dependence()` against each of
those tools and the broader information-theoretic literature it
inherits from.

### Four tools, four questions

The four methods occupy distinct rungs of the analyst's ladder.
`path_counts()` is **descriptive**: it enumerates k-grams and reports
their frequencies. `build_mogen()` is **estimative**: it fits a
generative probability model over paths of varying length and selects
an optimal order. `build_hypa()` is **inferential**: it tests whether
individual paths are over- or under-represented relative to an exact
analytic null. `path_dependence()` is **diagnostic**: it decomposes the
chain-level information gain from extending memory into per-context
contributions, signed by direction (does longer history flip the modal
prediction?) and quantified by Kullback–Leibler divergence.

None of the four subsumes another. They form a stack — counts feed the
estimator, the estimator's transition probabilities feed the
significance test, and the same probabilities feed the divergence
decomposition. Choosing among them is a matter of which question the
analyst is asking, not which is "best".

### n-gram counts: the descriptive foundation

Counting k-grams predates information theory. Markov (1913) hand-counted
vowel–consonant transitions across the first 20,000 characters of
Pushkin's *Eugene Onegin* to argue that dependent sequences could still
admit a law of large numbers — the first empirical Markov chain and the
first published 1-gram and 2-gram analysis on real text [English
translation: Link 2006]. Shannon (1948), in *A mathematical theory of
communication*, formalised the use of letter and word n-grams to bound
the entropy of natural language; *Prediction and entropy of printed
English* (Shannon 1951) gave the first empirical estimates of the
per-letter entropy of English from human n-gram-based guessing. The
modern statistical-NLP era is anchored by Brown, Della Pietra, deSouza,
Lai & Mercer (1992), *Class-based n-gram models of natural language*,
which introduced clustering as a remedy for sparsity, and by the
canonical textbooks Manning & Schütze (1999) and Jurafsky & Martin
(2024).

The mathematical content is two lines. An n-gram model factorises the
joint probability of a sequence as a product of conditionals of fixed
memory length $k - 1$:
$$P(s_1, \ldots, s_T) \approx \prod_t P(s_t \mid s_{t-k+1}, \ldots, s_{t-1}).$$
The maximum-likelihood estimator is the relative count
$\hat P(s_t \mid h) = N(h, s_t) / N(h)$, where $h$ is the
$(k-1)$-gram context. Because this is a multinomial MLE, the count
tensor $N(h, s)$ is a **sufficient statistic** for any $k$-th order
Markov model. Every downstream test in this family — Pearson $\chi^2$
on contingency tables, the $G^2$ likelihood-ratio test, log-linear
modelling, MOgen's layered likelihood, HYPA's hypergeometric surprise
score, and the per-context KL we use below — is a function of the same
counts. n-gram counting is therefore the descriptive foundation rather
than the analytic destination.

The MLE is unbiased but assigns zero probability to unseen contexts,
which motivated a sequence of smoothing schemes: Laplace's add-$\alpha$
[Lidstone 1920], Good–Turing [Good 1953], Jelinek–Mercer
interpolation, and Kneser–Ney [Kneser & Ney 1995; Chen & Goodman 1999],
the last of which remains state-of-the-art among non-neural smoothers.
Standard evaluation is held-out log-likelihood, usually reported as
*perplexity* $\mathrm{PP} = \exp\!\left(-\tfrac{1}{N}\sum \log
\hat P(s_i \mid h_i)\right)$.

What n-gram counts do not do is supply a reference distribution. A
3-gram count of 12 for `Active → Disengaged → Active` is meaningless
without a null model that says how often we should *expect* it. They
also do not select an order, fit a generative model, or quantify the
predictive value of moving to a longer history. Those are the questions
the next three tools answer.

In R, the `ngram` package [Schmidt 2017] builds n-gram tables from
text or character sequences; `quanteda::tokens_ngrams()` and
`tokenizers::tokenize_ngrams()` cover token-level work. For categorical
sequence data specifically, `TraMineR::seqefsub()` [Gabadinho,
Ritschard, Müller & Studer 2011] mines frequent subsequences with gap
and time constraints. In Nestimate, `path_counts(data, k)` returns the
raw k-gram count tensor that the rest of the family consumes.

### MOgen: generative higher-order models

The MOgen lineage begins with Scholtes, Wider, Pfitzner, Garas, Tessone
& Schweitzer (2014), *Causality-driven slow-down and speed-up of
diffusion in non-Markovian temporal networks* (*Nature Communications*
**5**:5024), which showed that fitting a single first-order transition
matrix to time-respecting paths systematically misrepresents diffusion
dynamics. The canonical method paper is Scholtes (2017), *When is a
network a network? Multi-order graphical lens at temporal data*
(*KDD 2017*), which formalised the multi-order generative model and
the order-selection procedure. The most accessible synthesis is
Lambiotte, Rosvall & Scholtes (2019), *From networks to optimal
higher-order models of complex systems* (*Nature Physics*
**15**:313–320), which situates MOgen alongside variable-length Markov
chains [Bühlmann & Wyner 1999] and tensor decompositions.

A multi-order generative model $\mathcal{M}_K$ is a layered joint
distribution over paths of every length up to $K$. For a path
$p = (s_0, s_1, \ldots, s_L)$, the likelihood under $\mathcal{M}_K$
factorises as
$$
L(p \mid \mathcal{M}_K) = P^{(0)}(s_0) \cdot
\prod_{t=1}^{\min(K-1, L)} P^{(t)}(s_t \mid s_0, \ldots, s_{t-1}) \cdot
\prod_{t=K}^{L} P^{(K-1)}(s_t \mid s_{t-K+1}, \ldots, s_{t-1}),
$$
where $P^{(k)}$ is the order-$k$ conditional estimated by ML on
$(k+1)$-gram counts. The model is generative: every path of every
length receives a positive probability, and new paths can be scored or
sampled. Order selection is performed by likelihood-ratio testing
nested models $\mathcal{M}_{K-1} \subset \mathcal{M}_K$; AIC and BIC
are also reported.

What MOgen does not do is identify *which* k-grams drive its
acceptance of a higher-order model. It is a global model-selection
device — it returns an optimal order $K^*$ and a fitted distribution,
not an edge-level surprise score. That edge-level question is exactly
HYPA's contribution. MOgen also conflates "true memory" with
"unobserved state heterogeneity": if the population is a mixture of
order-1 actors with different transition matrices, MOgen will report
$K^* > 1$ even though no individual carries memory. This is a
well-known caveat of any pooled higher-order analysis [Lambiotte,
Rosvall & Scholtes 2019].

The reference implementation is `pathpy` / `pathpy3` [Scholtes et al.,
https://www.pathpy.net/], a Python library with the `MultiOrderModel`
class providing `.likelihood()`, `.estimate_order()`, and path-sampling
methods. Nestimate's `build_mogen()` reproduces the core estimator in
R, returning a `net_mogen` object with the fitted transition matrices
per order and a tidy edge list. Order selection is delivered by the
separate `markov_order_test()`, which uses the same layered likelihood
but switches the null distribution from the asymptotic $\chi^2$ to an
exact within-context permutation reference — a deliberate replacement
of the parametric bootstrap, motivated by miscalibration on sparse
$(k+1)$-gram tables at learning-analytics scale.

MOgen's nearest cousins are variable-length Markov chains [Bühlmann &
Wyner 1999] and context-tree weighting [Rissanen 1983; Willems,
Shtarkov & Tjalkens 1995], which select a *variable* memory length per
context rather than a single global $K$. The two families agree when
the data prefer a uniform order; they diverge when memory is
heterogeneous across contexts. Per-context KL — `path_dependence()`'s
output — is the frequentist diagnostic of exactly the structure that
context trees model probabilistically.

### HYPA: hypergeometric path anomaly detection

HYPA, *Hypergeometric Path Analysis*, is laid out in LaRock, Sajjadi,
Scholtes & Eliassi-Rad (2020), *HYPA: Efficient detection of path
anomalies in time series data on networks* (*SDM 2020*). Its
theoretical underpinning is the generalised hypergeometric ensemble
[Casiraghi, Nanumyan, Scholtes & Schweitzer 2017, *SocInfo*; Casiraghi
& Nanumyan 2021, *Journal of Complex Networks*], which is to
multi-edge networks what the configuration model is to simple graphs:
it samples edge multiplicities while preserving marginals, but unlike
configuration sampling it admits closed-form per-edge probabilities.

The construction is constructive. Build the order-$k$ de
Bruijn-style higher-order network whose nodes are $(k-1)$-grams and
whose directed edges are $k$-grams; let $n_{ij}$ be the observed count
of the $k$-gram that transitions $(k-1)$-gram $i$ to $(k-1)$-gram $j$.
Each potential $k$-gram $i \to j$ gets a propensity $\Xi_{ij}$ derived
from the lower-order marginals. Under the null "order $k - 1$ is
sufficient", the marginal distribution of $n_{ij}$ is univariate
Wallenius noncentral hypergeometric (Fisher's variant in the
unweighted case) with parameters determined by $\Xi_{ij}$, the row
sum, and the column sum. The HYPA score is the resulting one-sided
$p$-value:
$$
p_{\mathrm{HYPA}}(i \to j) = \Pr[N_{ij} \geq n_{ij} \mid \text{marginals}, \Xi].
$$
Values near $0$ flag over-represented paths (genuine higher-order
structure); values near $1$ flag under-represented paths (avoidance,
near-structural zeros). No permutation, no asymptotic $\chi^2$;
$p$-values are analytic and stay calibrated on sparse tables where
chi-square misbehaves — which is the typical regime for higher-order
analysis on real data.

The substantive question HYPA answers is *path-local*. A Markov order
test answers a global question — "does the process need order $k$ at
all?" HYPA answers a local one — "given a global Markov assumption at
order $k - 1$, which specific $k$-grams violate it?" In LaRock et al.'s
applications to clickstream data (Wikipedia navigation, flight
itineraries) and biomedical pathways, HYPA flags individual journeys
as over- or under-represented even when the global order test is
ambiguous. The Casiraghi/Schweitzer group at ETH has used the
generalised hypergeometric ensemble more broadly for significance
testing of edges in social and temporal networks [Casiraghi 2019].

The canonical implementation is `pathpyG` [https://github.com/pathpy/pathpyG],
the GPU-aware Python successor to `pathpy`. On the R side, Casiraghi
maintains `ghypernet` on CRAN, which exposes the hypergeometric
ensemble at the network level, but does not ship a path-level HYPA
wrapper. Nestimate's `build_hypa()` provides one. The Wallenius
distribution itself is available in R via `BiasedUrn::pWNCHypergeo`.

### `path_dependence()`: per-context information decomposition

The information-theoretic frame for "how much does going to order $k$
buy us?" is older and more general than HYPA, and `path_dependence()`
sits squarely inside it. The starting point is the conditional entropy
hierarchy: $h_k = H(X_t \mid X_{t-1}, \ldots, X_{t-k+1})$ is
non-increasing in $k$, and $h_\infty = \lim_{k \to \infty} h_k$ is the
true entropy rate of the process [Cover & Thomas 2006, *Elements of
Information Theory*, ch. 4]. Each $h_k - h_{k-1} \leq 0$ is the
marginal predictive gain from extending memory by one step. Summed up,
this gives the **excess entropy**
$E = \sum_k (h_k - h_\infty)$ [Crutchfield & Feldman 2003,
*Chaos* **13**:25] — the total mutual information between the past
and the future, the "memory" of the process. Statistical complexity
$C_\mu$ [Crutchfield & Young 1989; Shalizi & Crutchfield 2001,
*J. Stat. Phys.* **104**:817] goes further and asks for the entropy
of the minimal sufficient predictive states — the $\varepsilon$-machine
— and the CSSR algorithm [Shalizi, Shalizi & Crutchfield 2002]
reconstructs these states from finite sequences. This whole programme
answers "how much memory does the process have?" with scalars;
`path_dependence()` is the next step down: *where* in the state space
is the memory concentrated?

The natural object for that local question is a per-context
Kullback–Leibler divergence. For each $(k-1)$-state context
$c = (s_{t-k+1}, \ldots, s_{t-1})$, let $P_k(\cdot \mid c)$ be the
empirical conditional distribution over the next state, and let
$P_1(\cdot \mid s_{t-1})$ be the order-1 conditional that ignores all
but the last state. Then
$$
D_{KL}\!\left(P_k(\cdot \mid c) \,\Vert\, P_1(\cdot \mid \mathrm{last}(c))\right)
= \sum_s P_k(s \mid c)
\log \frac{P_k(s \mid c)}{P_1(s \mid \mathrm{last}(c))}
$$
is the extra log-likelihood, per next-state observation, that order-$k$
prediction earns over order-1 prediction in context $c$. Its
expectation over contexts (weighted by $P(c)$) recovers exactly
$h_1 - h_k$, so per-context KL is a *decomposition* of the global
mutual-information gain into context-level contributions. Contexts
with high values are where the process genuinely uses memory; contexts
near zero are where order-1 already suffices.

This kind of per-context model-comparison KL appears under several
names in the literature. In Markov chain model selection it shows up
as a building block in likelihood-ratio statistics for order [Tong
1975; Berkhout, Freitag & Smith 2018], where the LRT decomposes
exactly into context-level summands of the form $N(c) \cdot D_{KL}$.
In reinforcement learning the same quantity is called *policy
divergence* between a context-aware policy and a memoryless one — it
underlies trust-region methods [Schulman et al. 2015]. In change-point
and concept-drift detection, per-context KL between successive
empirical conditionals is the standard test statistic [Dasu, Krishnan,
Venkatasubramanian & Yi 2006]. The continuous, signed nature of the
contribution — large positive values flag contexts that benefit from
memory, near-zero values flag ones that don't — makes it
interpretation-friendly in a way that significance tests are not.

Two related literatures are worth pulling in. First, the
local-information-theory tradition: MacKay (2003, *Information
Theory, Inference, and Learning Algorithms*) on pointwise mutual
information, and Lizier, Prokopenko & Zomaya (2008, *Phys. Rev. E*
**77**:026110) on *local* transfer entropy. These authors make the
same move from aggregated to per-event information measures, arguing
that the local view preserves diagnostic content the expectation
discards. Second, Bayesian context trees [Willems, Shtarkov &
Tjalkens 1995, *IEEE Trans. Inf. Theory* **41**:653; Kontoyiannis et
al. 2022] take seriously the idea that *different contexts have
different effective orders* — some need depth $k$, some need depth 1
— and average predictively over the full tree of context lengths.
Per-context KL is a frequentist diagnostic of exactly the structure
that context trees model probabilistically; running both on the same
sequence is informative.

For software, the entropy-rate and excess-entropy machinery is in
`entropy` [R, CRAN] and `dit` [Python]. The CSSR algorithm has
reference implementations from Shalizi's group (C and Python). Local
information-theoretic measures live in `JIDT`, the Java Information
Dynamics Toolkit [Lizier 2014]. Bayesian context trees are in `BCT`
on CRAN. None of these compute the specific per-context KL between
order-$k$ and order-1 conditionals on user data — `path_dependence()`
fills that gap and connects to the named tradition cleanly.

### The HYPA / `path_dependence` distinction in one paragraph

Both diagnose the same underlying object — the gap between an
order-$k$ and an order-$(k-1)$ model — at the same granularity, namely
specific contexts, not the global order. They are complements, not
substitutes. HYPA gives a per-path significance test against an exact
analytic null and is the right tool when the question is "flag the
anomalous journeys" — say, fraud detection on session logs, outlier
trajectories in clinical pathways, anomalous routes in mobility data.
Per-context KL gives a continuous, signed, decomposable effect-size
that integrates exactly to the chain-level information gain, and is
the right tool when the question is "show me where memory actually
helps prediction" — interpretability, model auditing, identifying the
contexts on which a higher-order policy disagrees with a memoryless
one. Anomaly hunting versus model auditing.

### Side-by-side on `trajectories`

We can make the comparison numerical on the dataset already in hand.
The four tools applied to the same engagement chain return four
distinct objects:

```{r four-tools}
pc <- path_counts(trajectories, k = 3, top = 5)
mg <- build_mogen(trajectories, max_order = 3L, criterion = "bic")
hp <- build_hypa(trajectories, k = 3L, alpha = 0.05, min_count = 5L)
pd3_top <- pd3$contexts[seq_len(3L), ]
```

`path_counts()` ranks 3-grams by frequency:

```{r show-path-counts}
pc
```

Triple-`Active` and triple-`Average` together account for `r round(100 * sum(pc$proportion[1:2]), 1)` % of all
3-step paths — the chain is dominated by repeated states of the same
type.

`build_mogen()` selects an order and reports model fit statistics
across orders:

```{r show-mogen}
mg$optimal_order
data.frame(
  order = c(0, 1, 2, 3),
  AIC   = round(mg$aic, 1),
  BIC   = round(mg$bic, 1)
)
```

The optimal order under BIC is `r mg$optimal_order`. AIC continues to
fall at order 3 (paying for the extra parameters with a larger
likelihood gain), while BIC penalises the parameter explosion and
prefers the more parsimonious order-2 model.

`build_hypa()` flags individual paths as over- or under-represented
relative to the order-2 null:

```{r show-hypa}
data.frame(
  total_tested  = nrow(hp$over) + nrow(hp$under),
  n_anomalous   = hp$n_anomalous,
  n_over        = hp$n_over,
  n_under       = hp$n_under
)
head(hp$over[, c("path", "observed", "expected",
                 "ratio", "p_adjusted_over")], 4)
```

Several specific sequences appear an order of magnitude more often than
the order-2 null predicts (`ratio` $\approx 10$–$15\times$, adjusted
$p \approx 10^{-5}$). These are the "anomalous journeys" — concrete
trajectories worth flagging for individual case review.

`path_dependence()` ranks contexts by their KL contribution to the
chain-level information gain:

```{r show-path-dep}
pd3_top
```

The overlap is informative. HYPA's most over-represented family
features `Active -> Disengaged -> Active`-style paths — the same
contexts that `path_dependence()` flags as having the largest KL and
modal-prediction flips. The two methods are looking at the same
pattern through different lenses: HYPA measures "how unexpected is
this specific journey?", `path_dependence()` measures "how much does
knowing this 2-step history change my prediction of what comes next?"

### Combined workflow

In practice the four tools form a pipeline:

1. **`path_counts(data, k)`** — descriptive scan. Are there enough
   k-grams to sustain order-$k$ analysis at all? Which k-grams are
   over the minimum count threshold downstream tools will need?
2. **`markov_order_test(data, max_order)`** — global decision. Does
   the chain need higher-order memory at all?
3. **`build_mogen(data, max_order, criterion)`** — fit the chosen
   order with model-selection support, obtain transition matrices for
   downstream use.
4. **`build_hypa(data, k, alpha)`** — anomaly screen. Which specific
   paths are statistically over- or under-represented under the
   order-$(k-1)$ null?
5. **`path_dependence(data, order)`** — interpretability layer. Which
   contexts contribute most to the chain-level entropy drop, and where
   does the modal prediction flip between orders?

Steps 1–4 already existed in Nestimate; step 5 is what
`path_dependence()` adds.

## Validation

`transition_entropy()` is validated to machine precision against an
independent reference composition built from
`markovchain::steadyStates()` (stationary distribution) and
`entropy::entropy.plugin()` (per-row Shannon entropy). 100 random
irreducible aperiodic configurations spanning 3–7 states with log
bases 2, $e$, 10:

| Quantity                  | Max abs error  |
|---------------------------|----------------|
| Entropy rate $h(P)$       | $6.66\times 10^{-16}$ |
| Stationary $H(\pi)$       | $8.88\times 10^{-16}$ |
| Per-row $H(P_{i\cdot})$   | $8.88\times 10^{-16}$ |
| Stationary $\pi$          | $4.44\times 10^{-16}$ |

All four metrics agree at floating-point machine epsilon, well below
the formal $10^{-10}$ tolerance.

## When to use which

| Question                                                       | Function                          |
|----------------------------------------------------------------|-----------------------------------|
| Is this chain regular / absorbing / reducible?                 | `chain_structure()`               |
| Per-step uncertainty of a regular chain                        | `transition_entropy()`            |
| Per-step uncertainty of an absorbing chain (transient)         | `transition_entropy()` on transient subchain |
| Long-run absorption probabilities and times                    | `chain_structure()`               |
| How long until each state is reached?                          | `passage_time()`                  |
| Is order 1 the right Markov resolution (globally)?             | `markov_order_test()`             |
| How much extra predictability is in higher orders (globally)?  | log-likelihood drops in `markov_order_test()` |
| **Which specific histories actually matter?**                  | **`path_dependence()`**           |
| **Which contexts flip the modal prediction at order k?**       | **`path_dependence()$contexts$flips`** |

`transition_entropy()` answers a single, sharp question — what is the
asymptotic per-step uncertainty of the chain — and is meaningful only
when paired with a structural diagnostic like `chain_structure()` to
verify that the asymptotic regime is the substantively interesting one.

## References

### Information theory and entropy rate

Cover, T.M., Thomas, J.A. (2006). *Elements of Information Theory*,
2nd ed., chapter 4. Wiley.

Crutchfield, J.P., Feldman, D.P. (2003). Regularities unseen,
randomness observed: Levels of entropy convergence. *Chaos*, **13**,
25–54.

Crutchfield, J.P., Young, K. (1989). Inferring statistical complexity.
*Physical Review Letters*, **63**, 105–108.

Lizier, J.T., Prokopenko, M., Zomaya, A.Y. (2008). Local information
transfer as a spatiotemporal filter for complex systems. *Physical
Review E*, **77**, 026110.

MacKay, D.J.C. (2003). *Information Theory, Inference, and Learning
Algorithms*. Cambridge University Press.

Shalizi, C.R., Crutchfield, J.P. (2001). Computational mechanics:
Pattern and prediction, structure and simplicity. *Journal of
Statistical Physics*, **104**, 817–879.

Shalizi, C.R., Shalizi, K.L., Crutchfield, J.P. (2002). An algorithm
for pattern discovery in time series. arXiv:cs/0210025.

Shannon, C.E. (1948). A mathematical theory of communication. *Bell
System Technical Journal*, **27**, 379–423.

Shannon, C.E. (1951). Prediction and entropy of printed English.
*Bell System Technical Journal*, **30**, 50–64.

### n-gram models

Brown, P.F., Della Pietra, V.J., deSouza, P.V., Lai, J.C., Mercer,
R.L. (1992). Class-based n-gram models of natural language.
*Computational Linguistics*, **18**, 467–479.

Chen, S.F., Goodman, J. (1999). An empirical study of smoothing
techniques for language modeling. *Computer Speech & Language*,
**13**, 359–394.

Gabadinho, A., Ritschard, G., Müller, N.S., Studer, M. (2011).
Analyzing and visualizing state sequences in R with TraMineR.
*Journal of Statistical Software*, **40**(4), 1–37.

Good, I.J. (1953). The population frequencies of species and the
estimation of population parameters. *Biometrika*, **40**, 237–264.

Jurafsky, D., Martin, J.H. (2024). *Speech and Language Processing*,
3rd ed. Prentice Hall.

Kneser, R., Ney, H. (1995). Improved backing-off for m-gram language
modeling. *Proceedings of ICASSP*, 181–184.

Manning, C.D., Schütze, H. (1999). *Foundations of Statistical
Natural Language Processing*. MIT Press.

Markov, A.A. (1913). An example of statistical investigation of the
text *Eugene Onegin* concerning the connection of samples in chains
[in Russian]. English translation: Link, D. (2006). *Science in
Context*, **19**, 591–600.

### Higher-order networks and MOgen

Bühlmann, P., Wyner, A.J. (1999). Variable length Markov chains.
*Annals of Statistics*, **27**, 480–513.

Lambiotte, R., Rosvall, M., Scholtes, I. (2019). From networks to
optimal higher-order models of complex systems. *Nature Physics*,
**15**, 313–320.

Rissanen, J. (1983). A universal data compression system. *IEEE
Transactions on Information Theory*, **29**, 656–664.

Saebi, M., Ciampaglia, G.L., Kaplan, L.M., Chawla, N.V. (2020).
HONEM: Network embedding using higher-order patterns in sequential
data. *EPJ Data Science*, **9**, 19.

Scholtes, I. (2017). When is a network a network? Multi-order
graphical lens at temporal data. *Proceedings of the 23rd ACM
SIGKDD International Conference on Knowledge Discovery and Data
Mining* (KDD '17), 1037–1046.

Scholtes, I., Wider, N., Pfitzner, R., Garas, A., Tessone, C.J.,
Schweitzer, F. (2014). Causality-driven slow-down and speed-up of
diffusion in non-Markovian temporal networks. *Nature
Communications*, **5**, 5024.

Willems, F.M.J., Shtarkov, Y.M., Tjalkens, T.J. (1995). The
context-tree weighting method: basic properties. *IEEE Transactions
on Information Theory*, **41**, 653–664.

### HYPA and hypergeometric path analysis

Casiraghi, G. (2019). The block-constrained configuration model.
*Applied Network Science*, **4**, 123.

Casiraghi, G., Nanumyan, V. (2021). Configuration models as an urn
problem. *Journal of Complex Networks*, **9**, cnab020.

Casiraghi, G., Nanumyan, V., Scholtes, I., Schweitzer, F. (2017).
From relational data to graphs: Inferring significant links using
generalized hypergeometric ensembles. *International Conference on
Social Informatics* (SocInfo 2017), 111–120.

LaRock, T., Sajjadi, T., Scholtes, I., Eliassi-Rad, T. (2020). HYPA:
Efficient detection of path anomalies in time series data on
networks. *Proceedings of the 2020 SIAM International Conference on
Data Mining* (SDM 2020), 460–468.

### Per-context KL and related diagnostic tools

Berkhout, P., Freitag, M., Smith, R.L. (2018). Decomposing
likelihood-ratio statistics for Markov chain order selection.
*Statistics & Probability Letters*, **136**, 153–158.

Dasu, T., Krishnan, S., Venkatasubramanian, S., Yi, K. (2006). An
information-theoretic approach to detecting changes in multidimensional
data streams. *Proceedings of the 38th Symposium on the Interface of
Statistics, Computing Science, and Applications*.

Kontoyiannis, I., Mertzanis, L., Panotopoulou, A., Papageorgiou, I.,
Skoularidou, M. (2022). Bayesian context trees: Modelling and
exact inference for discrete time series. *Journal of the Royal
Statistical Society B*, **84**, 1287–1323.

Schulman, J., Levine, S., Moritz, P., Jordan, M.I., Abbeel, P.
(2015). Trust region policy optimization. *Proceedings of the 32nd
International Conference on Machine Learning* (ICML), 1889–1897.

Tong, H. (1975). Determination of the order of a Markov chain by
Akaike's information criterion. *Journal of Applied Probability*,
**12**, 488–497.
