---
title: "Tutorial: Qualitative structure of a discrete-time Markov chain"
subtitle: "When is the stationary distribution meaningful? When can a learner get from A to B? When do they get stuck?"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  out.width  = "100%",
  fig.width  = 8,
  fig.height = 5,
  dpi        = 96,
  warning    = FALSE,
  message    = FALSE
)
library(Nestimate)
```

## What this tutorial answers

`passage_time()` and `markov_stability()` both return a stationary
distribution and use it as the basis for every per-state metric they report.
That is only meaningful if the underlying chain has *one* long-run distribution
to converge to — which is to say, if the chain is **regular** (irreducible
plus aperiodic). When it is not, the reported `stationary_prob` is averaging
across distinct behavioural phases that the learner never actually mixes
between. `chain_structure()` is the diagnostic that tells you which case
you are in.

It also answers two interpretive questions that come up every time a coded
sequence dataset has a clear terminal behaviour ("submitted", "abandoned",
"gave up"): which states *funnel* into which, and how long does the funnel
take.

This tutorial walks through `chain_structure()` on four bundled datasets:

- `trajectories` — engagement states (Active / Average / Disengaged).
- `group_regulation_long` — collaborative-regulation actions across 27,533
  events, six courses, two achievement strata.
- `ai_long` — AI-coded vibe-coding interaction sessions.
- `human_long` — the same sessions coded by a human rater.

Every section follows the same shape: state the question, run a single
`Nestimate` call, read the printed result, draw a conclusion.

---

## What `chain_structure()` returns

```{r}
P <- matrix(c(
  0.5, 0.3, 0.2,
  0.4, 0.4, 0.2,
  0.0, 0.0, 1.0
), nrow = 3, byrow = TRUE,
   dimnames = list(c("Working", "Distracted", "GivenUp"),
                   c("Working", "Distracted", "GivenUp")))
chain_structure(P)
```

The header summarises four chain-level questions: is the chain
irreducible (one communicating class)? aperiodic? regular (= both)? does
it satisfy detailed balance against its stationary distribution? Below
that, every state is classified as **absorbing**, **recurrent**, or
**transient**, and — when absorbing states exist — the absorption
probabilities and expected absorption times tell you what happens to the
transient states.

In this synthetic example, GivenUp is absorbing (`P[GivenUp, GivenUp] = 1`)
and the other two are transient: every learner eventually gives up,
taking five steps on average from either Working or Distracted. The
chain is *not* regular because there is no return path from GivenUp;
running `markov_stability(P)` here would report a stationary distribution
concentrated entirely on GivenUp, which is correct as a long-run statement
but uninformative about how long Working and Distracted persist before
absorption — that information lives in `mean_absorption_time`, not
`stationary_prob`.

The `plot()` method renders the hitting-probability matrix with rows
and columns ordered by communicating class, so the block structure of
the chain is visible at a glance. Axis labels are coloured by
classification (absorbing in vermillion, recurrent in blue, transient
in orange).

```{r, fig.height = 4}
plot(chain_structure(P))
```

Reading the heatmap: the **GivenUp** column is solid dark — every
state, including itself, reaches GivenUp with probability 1. The
**GivenUp** row is white except in the GivenUp column — once you are
in GivenUp, you cannot reach Working or Distracted. That asymmetry is
the visual signature of an absorbing state.

The remainder of the tutorial uses `chain_structure()` as a
*precondition check* on real data, then reads off the per-state
absorption and hitting structure where it applies.

---

## Engagement trajectories

The `trajectories` dataset records 138 students' engagement states across
15 weekly time points. The three states are coarse: Active, Average,
Disengaged. The first analytic question is whether transitions between
these states form a regular chain; if so, `markov_stability()`'s long-run
output is interpretable; if not, we need a different reading.

```{r}
net_traj <- build_network(as.data.frame(trajectories), method = "relative")
cs_traj  <- chain_structure(net_traj)
cs_traj
```

The chain is irreducible, aperiodic, regular, and not reversible. Every
state is recurrent — no engagement category is a one-way door. The
*non-reversibility* is the substantive finding here: it says students do
not flow symmetrically between the three states. There is a preferred
direction in the long-run flux, even though every state is recurrent.

Because the chain is regular, `markov_stability()` can be reported
without further hedging:

```{r}
markov_stability(net_traj)
```

Active and Average have similar return times (about 2.7 weeks); the
sojourn times — how long a student stays in a state once they enter it —
are all near 1 week, which means trajectories are noisy at the weekly
grain.

The hitting probabilities are saturated:

```{r, fig.height = 3.5}
plot(cs_traj)
```

Every state can reach every other with probability 1 — the heatmap is
saturated dark across all 9 cells. With three recurrent states and no
absorbing endpoint, the educational interpretation is straightforward:
students cycle between engagement levels for the duration of the
course, and the long-run average stationary distribution genuinely
describes the population.

---

## Group regulation actions

`group_regulation_long` contains 27,533 collaborative-regulation events
from 23 students across six courses, two achievement levels, and 14
distinct actions (cohesion, consensus, discuss, plan, monitor, etc.).
The first question, again, is whether the regulation chain is regular —
in particular, whether any regulation behaviour is a dead-end that
students never recover from.

```{r}
net_gr <- build_network(group_regulation_long,
                        action = "Action",
                        actor  = "Actor",
                        time   = "Time",
                        method = "relative")
cs_gr <- chain_structure(net_gr)
cs_gr
```

The chain is irreducible, aperiodic, regular, and *not* reversible.
Substantively, the regularity finding says every regulation action is
reachable from every other and there are no absorbing behaviours — if
some single action had been a dead-end (for example "disengage" or
"exit"), it would warrant a separate funnel analysis. The fact that
none is absorbing means `markov_stability()` reads directly as a
portrait of the population.

Non-reversibility is the substantively important finding. It says the
flux between any pair of regulation actions is asymmetric at
equilibrium: the rate of transitions from action *i* to *j* differs
from the rate from *j* to *i*. There *is* a preferred direction
through the regulation behaviours — collaborative regulation is a
sequenced process, not a memoryless mix of marginally-preferred
actions. The next question, then, is whether the *direction* of that
sequence differs between high and low achievers.

```{r}
net_gr_grp <- build_network(group_regulation_long,
                            action = "Action",
                            actor  = "Actor",
                            time   = "Time",
                            group  = "Achiever",
                            method = "relative")
cs_high <- chain_structure(net_gr_grp$High)
cs_low  <- chain_structure(net_gr_grp$Low)

data.frame(
  group         = c("High", "Low"),
  irreducible   = c(cs_high$is_irreducible, cs_low$is_irreducible),
  aperiodic     = c(cs_high$is_aperiodic,   cs_low$is_aperiodic),
  regular       = c(cs_high$is_regular,     cs_low$is_regular),
  reversible    = c(cs_high$is_reversible,  cs_low$is_reversible),
  n_classes     = c(length(cs_high$communicating_classes),
                    length(cs_low$communicating_classes)),
  stringsAsFactors = FALSE
)
```

Both strata produce regular non-reversible chains with one
communicating class — the qualitative chain structure is identical.
The substantive difference between high and low achievers, if any, is
therefore in the quantitative parameters (transition probabilities,
dwell times, stationary frequencies), not in the *qualitative* shape
of the regulation process. That is exactly the question
`passage_time()` on each group answers, and it is exactly the question
that would be *meaningless* if the two chains had different qualitative
structure.

```{r, fig.height = 5, fig.width = 7}
plot(cs_high, show_values = FALSE)
```

```{r, fig.height = 5, fig.width = 7}
plot(cs_low, show_values = FALSE)
```

The two heatmaps show a similar global pattern: every regulation
action is reachable from every other (no white off-block region),
which confirms the regularity finding numerically; the diagonal
intensity (return probability) varies across actions in both groups,
identifying which regulation behaviours are most self-reinforcing.

---

## Vibe-coding interactions: AI- vs human-coded

The `ai_long` and `human_long` datasets code 429 vibe-coding sessions
into a small set of cluster categories (three at the coarsest level used
here). The two coders disagree on the boundary of nearly every category,
but the question for `chain_structure()` is coarser: do they agree on
the *qualitative* structure of the resulting chain?

```{r}
net_ai    <- build_network(ai_long,    action = "cluster",
                           actor = "session_id", time = "code_order",
                           method = "relative")
net_human <- build_network(human_long, action = "cluster",
                           actor = "session_id", time = "code_order",
                           method = "relative")
cs_ai    <- chain_structure(net_ai)
cs_human <- chain_structure(net_human)

data.frame(
  coder       = c("AI", "Human"),
  n_states    = c(length(cs_ai$states), length(cs_human$states)),
  irreducible = c(cs_ai$is_irreducible, cs_human$is_irreducible),
  aperiodic   = c(cs_ai$is_aperiodic,   cs_human$is_aperiodic),
  regular     = c(cs_ai$is_regular,     cs_human$is_regular),
  reversible  = c(cs_ai$is_reversible,  cs_human$is_reversible),
  absorbing   = c(length(cs_ai$absorbing_states),
                  length(cs_human$absorbing_states)),
  stringsAsFactors = FALSE
)
```

The two coding schemes produce qualitatively identical chains: regular,
non-reversible, no absorbing states. So even though raters disagree on
which utterance maps to which category, they agree on the topological
shape of the conversational process. That is a stronger inter-rater
guarantee than category-level agreement statistics provide.

Non-reversibility for both says the same thing it said for
`trajectories`: there is a preferred direction in the long-run flow
between behaviours. Coding sessions move forward, not symmetrically.

For a per-state comparison, the cleanest readout is the diagonal of the
hitting-probability matrix. `H[i, i]` is the probability of *returning*
to state *i* on a path of length at least one — a measure of how
self-reinforcing each behaviour is.

```{r}
cmp <- data.frame(
  cluster   = cs_ai$states,
  H_self_ai = round(diag(cs_ai$hitting_probabilities), 3),
  stringsAsFactors = FALSE
)
cmp$H_self_human <- round(diag(cs_human$hitting_probabilities)[match(cmp$cluster, cs_human$states)], 3)
cmp$delta <- round(cmp$H_self_ai - cmp$H_self_human, 3)
cmp[order(-abs(cmp$delta)), ]
```

The largest cross-coder disagreement is in how often a behaviour
returns to itself — the `delta` column. Categories where the AI
coder reports markedly higher `H_self` than the human coder are
categories where the AI is *over-fragmenting* the conversation
(the same underlying behaviour, sliced into more frequent
self-transitions); categories where the human reports higher
`H_self` are where the human is over-aggregating. This is a more
diagnostic measure of coder disagreement than any category-frequency
statistic, because it captures the *temporal* signature of the
coding choice rather than just the marginal counts.

---

## When to use which

| Question | Function |
|---|---|
| Is this chain regular? Should I trust `markov_stability()`'s stationary output? | `chain_structure()$is_regular` |
| Is there a one-way pathway through behaviours? | `chain_structure()$is_reversible` (FALSE means yes) |
| Are any behaviours dead-ends? | `chain_structure()$absorbing_states` |
| Given a learner is in behaviour X, what is the probability of *eventually* reaching Y? | `chain_structure()$hitting_probabilities[X, Y]` |
| Given a learner is in behaviour X, expected number of steps to absorption? | `chain_structure()$mean_absorption_time[X]` |
| Given absorbing states exist, which one does each transient state most likely funnel into? | `chain_structure()$absorption_probabilities` |
| Long-run state frequencies *given* a regular chain | `markov_stability()$stationary_prob` |
| Mean steps from state *i* to state *j* *given* a regular chain | `passage_time()$matrix[i, j]` |

Run `chain_structure()` first; the output tells you whether the rest of
the long-run analysis is meaningful. If `is_regular = FALSE`, the
absorption-and-hitting analyses inside `chain_structure()` itself replace
the stationary-distribution-based summaries.

## References

- Kemeny, J. G., and Snell, J. L. (1976). *Finite Markov Chains*.
  Springer-Verlag. (Fundamental matrix, absorption-time formula.)
- Norris, J. R. (1997). *Markov Chains*. Cambridge University Press.
  (Hitting probabilities as the minimal non-negative solution of the
  recurrence; Theorem 1.3.2.)
- Spedicato, G. A. (2017). *Discrete Time Markov Chains with R*.
  R Journal 9(2), 84–104. (Reference implementation in the
  `markovchain` package; `chain_structure()` is validated to machine
  precision against `markovchain::*` across 100 random configurations
  in `local_testing_and_equivalence/test-equiv-chain_structure.R`.)
