---
title: "State Frequencies: Marimekko and Bar Plots"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 6, fig.height = 3.2, dpi = 96
)
```

The question this tutorial answers: *how do the relative frequencies of
states differ across groups?* `plot_state_frequencies()` is the
package's native generic for this pattern. It dispatches across
`netobject`, `netobject_group`, `mcml`, and `htna`. The rendering is
class-aware:

- For `mcml` and `htna`, the state set in each group/cluster is
  typically disjoint (different clusters carry different states; AI
  uses different actions than Human). The default is a per-facet
  layout: each panel is a stand-alone treemap with its **own** legend
  showing only that panel's states, all panels arranged side-by-side.
- For a single `netobject` (no groups) or a `netobject_group` (where
  groups share the same state set), the default is a single-chart
  layout with one shared legend at the bottom. Each group still gets
  its own treemap panel; only the legend is shared.

Both layouts use the colorblind-safe Okabe-Ito palette.

```{r}
library(Nestimate)
```

## A single network

A single `netobject` has no groups, so the chart is a single
squarified treemap: each state becomes a rectangle whose area is
exactly proportional to its overall share.

```{r single-net}
data(trajectories, package = "Nestimate")
nw <- build_network(as.data.frame(trajectories),
                    method = "relative", format = "wide")

plot_state_frequencies(nw)
```

`Average` covers roughly two-thirds of the panel, occupying the
largest tile; `Active` and `Disengaged` split the remainder into
smaller tiles. Reading areas directly: tile size = proportion.

The inline tile labels are controlled via `label`. The default is
`"prop"` (proportions). To show raw counts instead, set
`label = "freq"`; for both, set `label = "both"`; to suppress
inline labels entirely, `label = "none"`.

```{r single-both}
plot_state_frequencies(nw, label = "both")
```

The bar style produces horizontal bars sorted by frequency:

```{r single-bars}
plot_state_frequencies(nw, style = "bars")
```

For bars, `metric` controls whether the bar length encodes the
proportion (default) or the raw count:

```{r single-bars-freq}
plot_state_frequencies(nw, style = "bars", metric = "freq",
                       label = "freq")
```

## Per-group networks (faceted treemaps)

`build_network()` called with `group =` returns a `netobject_group`.
Each group becomes its own treemap panel --- the within-group state
mix is read directly from tile areas.

```{r group-net}
data(group_regulation_long, package = "Nestimate")
nw_g <- build_network(group_regulation_long,
                      method = "relative", format = "long",
                      actor = "Actor", action = "Action",
                      order = "Time", group = "Course")

plot_state_frequencies(nw_g)
```

Three panels, one per course. Each panel is a squarified treemap of
that course's action distribution: the largest tile is the
most-frequent action in that course. Cross-panel comparison is by
which colored tile dominates each panel.

The bar style switches to horizontal bars, faceted by group:

```{r group-bars}
plot_state_frequencies(nw_g, style = "bars")
```

## Multi-cluster multi-level (MCML)

`build_mcml()` partitions the state set into clusters. Because the
state vocabularies in different clusters are typically disjoint, the
default is a per-cluster layout: each cluster becomes its own
treemap panel with its own legend listing only that cluster's states.

```{r mcml-net, fig.width = 8, fig.height = 8}
Interactions <- rio::import(
  "/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/complicated_nw2.csv"
)
Nodesin <- rio::import(
  "/Users/mohammedsaqr/Downloads/mcml_code/Mcdata/Codemapping.xlsx"
)

clusters_list <- split(Nodesin$name, Nodesin$groups)
labels_df     <- Nodesin[, c("name", "Code")]

mc <- build_mcml(
  Interactions,
  clusters = clusters_list,
  actor    = "course_group",
  time     = "timestamp",
  action   = "From",
  type     = "tna",
  labels   = labels_df
)
plot_state_frequencies(mc, label = "all", ncol = 2)
```

One stand-alone treemap per cluster, each with its own legend listing
only that cluster's codes. Each panel restarts the palette from the
first color, so within-cluster ordering is read by tile size; cross-
panel comparison is by cluster width and within-cluster spread.

## Heterogeneous networks (htna)

An `htna` object carries a node-level grouping (`$node_groups`) that
labels each state by an actor or context. Each group becomes its own
marimekko panel.

```{r htna-net, eval = requireNamespace("htna", quietly = TRUE)}
library(htna)
data(ai_long, package = "Nestimate")
data(human_long, package = "Nestimate")

ht <- build_htna(list(AI = ai_long, Human = human_long),
                 action = "code", session = "session_id",
                 order = "code_order")

plot_state_frequencies(ht)
```

Two stand-alone treemaps side-by-side --- AI on the left with its
own legend (only AI's actions), Human on the right with its own
legend (only Human's actions). Each legend is enclosed in a thin
border, sitting next to its panel like a key card. Colors stay
consistent across panels so the same code keeps the same color.

## The `plot_mosaic()` primitive

The hierarchical-marimekko geometry is also exposed as a low-level
primitive. Any tidy `data.frame(group, state, weight)` works:

```{r mosaic-primitive}
df <- data.frame(
  group  = rep(c("A", "B", "C"), each = 3),
  state  = rep(c("s1", "s2", "s3"), 3),
  count  = c(10, 5, 2,  7, 8, 3,  4, 6, 12)
)
plot_mosaic(df, x = "group", y = "state", weight = "count")
```

The `fill` argument can map to any other column when residual or
effect-size shading is wanted instead of state colors. This makes
`plot_mosaic()` reusable for residual-style mosaic plots downstream
(for example, in permutation-test visualisations).

## Label and legend combinations

Three arguments control all annotation:

- `label` --- text inside each tile/bar.
  `"prop"` (default), `"freq"`, `"both"`, `"state"`, `"all"`, `"none"`.
- `legend` --- where the state legend sits.
  `"bottom"` (default), `"top"`, `"right"`, `"left"`, `"none"`.
- `legend_frame` --- `"none"` (default) or `"border"` for an enclosed legend.

Below: a gallery on the same `netobject_group` showing each combination.

### Default (proportions, legend at bottom)

```{r combo-default}
plot_state_frequencies(nw_g)
```

Tile size = state share; percentage labels on the top tiles; legend
names the colors at the bottom. The same state always lands in the
same relative position across facets, so cross-panel comparison is by
*size of the same colored tile*.

### Counts in tiles, legend at right (vertical)

```{r combo-freq-right}
plot_state_frequencies(nw_g, label = "freq", legend = "right")
```

Right-side legend saves vertical space; useful when the figure is
tall.

### Both numbers, legend at right with border

```{r combo-both-right-frame}
plot_state_frequencies(nw_g, label = "both", legend = "right",
                       legend_frame = "border")
```

`"1,234 (66%)"` inline on each tile; legend enclosed in a thin
rectangle (`legend_frame = "border"`) so it reads as a single boxed
unit beside the cards.

### Both numbers, no legend

```{r combo-both-nolegend}
plot_state_frequencies(nw_g, label = "both", legend = "none")
```

Counts + proportions inline, legend hidden because the tiles already
carry both numbers. Color is decorative here.

### State names in tiles, no legend (handout style)

```{r combo-state-nolegend}
plot_state_frequencies(nw_g, label = "state", legend = "none")
```

State names written directly on the tiles, no separate legend.
Useful for printable handouts. Small tiles drop their labels
automatically; tile text auto-rotates 90 degrees in tall-narrow
tiles to follow the tile's long axis.

### State name + proportion inline, framed legend

```{r combo-all-frame}
plot_state_frequencies(nw_g, label = "all", legend_frame = "border")
```

Highest information density: tiles read `"Average (66%)"` on a
single line; the legend is still present but enclosed in a border so
it reads as a key card.

### Legend on top, horizontal

```{r combo-legend-top}
plot_state_frequencies(nw_g, legend = "top")
```

Caption-style legend above the cards. Good for slide layouts where
the chart needs to anchor to the bottom of the frame.

### Legend on left, framed

```{r combo-legend-left-frame}
plot_state_frequencies(nw_g, legend = "left", legend_frame = "border")
```

Vertical legend on the left side, framed --- mirrors the right-side
boxed pattern.

### Frequencies only, no labels in tiles

```{r combo-prop-only-state}
plot_state_frequencies(nw_g, label = "none", legend = "right",
                       legend_frame = "border")
```

Pure tile-area encoding: no inline text at all, only color, with a
boxed right-side legend acting as the legend card.

## Label/legend on the htna treemap

```{r htna-combo, eval = requireNamespace("htna", quietly = TRUE)}
plot_state_frequencies(ht, label = "all", legend = "right",
                       legend_frame = "border")
```

State names + proportions written inline on each actor's tiles, with
a framed right-side legend providing the color key.

## Per-facet legends

For `htna` (one panel per actor) and `mcml` (one per cluster), the
states in each panel are often disjoint --- AI uses a different action
set than Human; the Engaged cluster contains different states than
Disengaged. A single global legend mixes them all together and is
hard to read.

`legend = "per_facet"` renders each panel as its own treemap with an
isolated legend that only lists the states present in that panel.
Colors stay consistent across panels (a state always maps to the same
color) so cross-panel comparison still works.

### htna with per-actor legends

```{r htna-per-facet, eval = requireNamespace("htna", quietly = TRUE) && requireNamespace("gridExtra", quietly = TRUE)}
plot_state_frequencies(ht, legend = "per_facet", label = "all")
```

Two panels, AI on the left and Human on the right, each with its own
legend showing only that actor's actions. Enclosed-square legend by
default for this mode.

### mcml with per-cluster legends

```{r mcml-per-facet, eval = requireNamespace("gridExtra", quietly = TRUE)}
plot_state_frequencies(mc, legend = "per_facet", label = "all")
```

Each cluster becomes a stand-alone treemap with a per-cluster legend.
Useful when the cluster state sets are very different.

### netobject_group with per-group legends

```{r group-per-facet, eval = requireNamespace("gridExtra", quietly = TRUE)}
plot_state_frequencies(nw_g, legend = "per_facet", label = "all")
```

For groups that share the same state set (every course has every
action), the per-facet legends will repeat the same set of states. In
that case a single shared legend (the default) reads cleaner. Per-
facet really shines when the state vocabularies differ across panels.

## Standardized-residual mosaic

`style = "mosaic"` answers a different question: *which states are
over- or under-represented in each group, relative to the marginal
distribution?* The chart is a state x group contingency table where:

- Row heights are proportional to the state's total frequency across
  all groups
- Column widths are proportional to the group's total count
- Each cell is colored by its **standardized residual**
  $r_{ij} = (O_{ij} - E_{ij}) / \sqrt{E_{ij}\,(1 - p_{i\cdot})(1 - p_{\cdot j})}$,
  matching the formula used in `sequence_compare()` --- blue for
  over-represented, red for under-represented, white for at-expected.

```{r mosaic-groupreg, fig.height = 4.16}
plot_state_frequencies(nw_g, style = "mosaic", label = "all")
```

`consensus` is sharply over-represented in course A (residual ~+7) and
sharply under-represented in B and C; `plan` shows the opposite
pattern. The reading rule: any cell darker than ±2 is statistically
non-trivial under the chi-square independence model.

## Bar combinations

```{r bars-prop-default}
plot_state_frequencies(nw_g, style = "bars")
```

```{r bars-freq-both}
plot_state_frequencies(nw_g, style = "bars", metric = "freq", label = "both")
```

```{r bars-prop-state-nolegend}
plot_state_frequencies(nw_g, style = "bars", label = "state",
                       legend = "none")
```

```{r bars-prop-all-frame}
plot_state_frequencies(nw_g, style = "bars", label = "all",
                       legend = "right", legend_frame = "border")
```

## When to use which

- For `mcml`, `plot_state_frequencies()` automatically renders the
  hierarchical 2D marimekko --- this is the right chart whenever
  states partition into clusters.
- For `netobject_group` and `htna`, the per-panel state-column
  marimekko is the default --- one panel per group, each showing
  variable-width state columns. Use `style = "bars"` when raw counts
  are the report unit.
- For a single `netobject` without groups, the marimekko collapses to
  a single panel --- easier to read than a stacked column.
- Reach for `plot_mosaic()` directly when the input is not a Nestimate
  object but the marimekko semantics fit (custom tabulations, residual
  fills, etc.).
