---
title: "Nestimate Function Demo"
format: html
---

```{r setup, message=FALSE, warning=FALSE}
library(Nestimate)
d <- read.csv("saqrvibcoding.csv", stringsAsFactors = FALSE)

```





# 2. Frequencies

```{r}

freqs <- build_network(d, method = "frequency", format = "long", action = "superclass")
cograph::splot(freqs$matrix)



reld <- build_network(d, method = "relative",  action = "superclass") 


as_tna(reld)

```

# 3. Group Networks

```{r}
grp <- build_network(wide, method = "relative", group = "Project")
print(grp)

```

# 4. Multilevel (between/within)

```{r}
net_ml <- build_network(wide, method = "relative", id = "Project",
                        level = "both")
print(net_ml)
round(net_ml$between$matrix, 3)
round(net_ml$within$matrix, 3)
```

# 5. Bootstrap

```{r}
boot <- bootstrap_network(net_rel, iter = 500, seed = 42)
print(boot)
summary(boot)
```

# 6. Permutation Test

```{r}
# Compare two large projects
p7 <- wide[wide$Project == "Project_7", grep("^T", names(wide))]
p5 <- wide[wide$Project == "Project_5", grep("^T", names(wide))]
net7 <- build_network(p7, method = "relative")
net5 <- build_network(p5, method = "relative")
perm <- permutation_test(net7, net5, iter = 500, seed = 42)
print(perm)
```

# 7. Reliability

```{r}
rel <- reliability(net_rel, iter = 200, seed = 42)
print(rel)
```

# 8. Centrality Stability

```{r}
cs <- centrality_stability(net_rel, iter = 200, seed = 42)
print(cs)
```

# 9. Clustering

```{r}
cl <- cluster_data(seq_wide, k = 3, seed = 42)
print(cl)
summary(cl)
plot(cl, type = "silhouette")
plot(cl, type = "mds")
```

## With covariates

```{r}
# Bin projects by size for a meaningful covariate
proj_sizes <- table(wide$Project)
wide$ProjSize <- ifelse(proj_sizes[wide$Project] <= 5, "Small",
                 ifelse(proj_sizes[wide$Project] <= 20, "Medium", "Large"))
cl_cov <- cluster_data(wide, k = 3, seed = 42, covariates = "ProjSize")
summary(cl_cov)
plot(cl_cov, type = "predictors")
```

## Per-cluster networks

```{r}
cl_nets <- build_network(cl)
print(cl_nets)
```

# 10. Mixed Markov Model

```{r}
mmm <- build_mmm(seq_wide, k = 2, n_starts = 10, seed = 42)
print(mmm)
summary(mmm)
```

## MMM with covariates

```{r}
mmm_cov <- build_mmm(wide, k = 2, n_starts = 10, seed = 42,
                      covariates = "ProjSize")
summary(mmm_cov)
```

## Model comparison

```{r}
comp <- compare_mmm(seq_wide, k = 2:4, n_starts = 5, seed = 42)
print(comp)
```

# 11. Higher-Order Networks

```{r}
seq_wide <- seq_wide

hon <- build_hon(seq_wide, max_order = 2)
print(hon)
head(hon$edges)

honem <- build_honem(hon, dim = 10)
print(honem)
head(honem$edges)

mogen <- build_mogen(seq_wide, max_order = 3)
print(mogen)
head(mogen$edges)

hypa <- build_hypa(seq_wide, k = 2)
print(hypa)
head(hypa$edges)
```

# 12. Pathways

```{r}
paths <- pathways(hon)
head(paths, 15)

pc <- path_counts(seq_wide, k = 3)
head(pc, 15)
```

# 13. Extract helpers

```{r}
extract_transition_matrix(net_rel)
head(extract_edges(net_rel))
extract_initial_probs(seq_wide)
```

# 14. Psychological Networks

```{r}
# Build per-session state frequency counts for association networks
freq_mat <- table(d$session_id, d$superclass)
freq_df <- as.data.frame.matrix(freq_mat)

net_cor <- build_network(freq_df, method = "cor")
round(net_cor$matrix, 3)

net_pcor <- build_network(freq_df, method = "pcor")
round(net_pcor$matrix, 3)

net_glasso <- build_network(freq_df, method = "glasso")
round(net_glasso$matrix, 3)
```

## Predictability

```{r}
pred <- predictability(net_cor)
print(pred)
```

## boot_glasso

```{r}
bg <- boot_glasso(net_glasso, iter = 200, seed = 42)
print(bg)
summary(bg)
```

# 15. Data Conversion

```{r}
long_back <- wide_to_long(seq_wide, time_prefix = "T")
head(long_back)
```

# 16. Custom Estimator

```{r}
register_estimator("binary", function(data, ...) {
  mat <- extract_transition_matrix(
    build_network(data, method = "relative", ...)
  )
  (mat > median(mat[mat > 0])) * 1
}, description = "Binary threshold", directed = TRUE)

net_bin <- build_network(seq_wide, method = "binary")
net_bin$matrix
remove_estimator("binary")
```

# 17. Window-based TNA

```{r}
# Convert to one-hot binary matrix
oh <- action_to_onehot(d, action_col = "superclass")
head(oh[, 1:8])

# Window-based TNA on one-hot data
w <- wtna(oh, window_size = 10, type = "relative",
          codes = unique(d$superclass))
print(w)
```

```{r}
sessionInfo()
```
