---
title: "Network Estimation and Visualization with Nestimate + cograph"
description: |
    From raw sequence data to publication-ready statistical network plots ---
    build, bootstrap, compare, cluster, and visualize using Nestimate and cograph together.
date: "2026-04-09"
author:
  - name: Mohammed Saqr
    url: https://saqr.me
    affiliation: University of Eastern Finland
  - name: Sonsoles López-Pernas
    url: https://sonsoles.me
    affiliation: University of Eastern Finland
categories:
  - tutorial
  - R
  - visualization
  - Nestimate
warning: false
message: false
fig.align: center
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    fig.width: 8
    fig.height: 7
    dpi: 300
---

```{r}
#| label: setup
#| include: false
library(Nestimate)
library(cograph)
```

# Introduction

**Nestimate** is a unified network estimation package for R. It provides a single entry point --- `build_network()` --- that accepts sequence data, wide-format behavioral data, or numeric matrices and estimates networks using any of 10 built-in methods. The result is a `netobject` that inherits `cograph_network`, so every object produced by Nestimate can be plotted with `splot()` directly.

The core design principle is that **estimation and visualization are separate concerns**: Nestimate handles all computation (network construction, bootstrapping, permutation testing, clustering, multi-cluster analysis), while cograph handles all rendering. This separation means the analyst writes estimation code once and gets publication-ready figures without adapter code or manual class conversion.

Beyond single-network estimation, Nestimate provides a complete statistical toolkit:

- **Bootstrap stability analysis** (`bootstrap_network`) --- identifies which edges survive resampling and which are sampling artifacts.
- **Permutation-based group comparison** (`permutation_test`) --- tests whether edges differ significantly between conditions.
- **Multi-cluster multi-level analysis** (`cluster_summary`, `build_mcml`) --- aggregates node-level networks into cluster-level summaries for hierarchical visualization.
- **Centrality stability** (`centrality_stability`) --- assesses whether centrality rankings are robust to case dropping.
- **Group-level operations** --- bootstrapping, permutation testing, and plotting work on grouped networks with no additional code.

This tutorial walks through the full pipeline with the `group_regulation_long` dataset shipped with Nestimate.

---

# Data

`group_regulation_long` is a long-format dataset of collaborative regulation behaviors coded across student groups. Each row is one coded action in a session, with columns identifying the actor, the group, achievement level, course, timestamp, and the regulation behavior.

```{r}
#| label: data
data("group_regulation_long", package = "Nestimate")
head(group_regulation_long)
```

The nine regulation behaviors are: `adapt`, `cohesion`, `consensus`, `coregulate`, `discuss`, `emotion`, `monitor`, `plan`, and `synthesis`. These form the nodes of the networks we estimate below.

---

# Estimating Networks

## A single network

`build_network()` estimates a network from raw data. The `method` argument selects the estimation algorithm. Here we use `"relative"`, which computes row-normalized transition probabilities --- the standard approach in transition network analysis (TNA). Unlike constructing a TNA model manually (preparing wide-format data, computing the transition matrix, normalizing), `build_network()` handles all data preparation internally.

```{r}
#| label: build-network
net <- build_network(
  group_regulation_long,
  action = "Action", actor = "Actor",
  method = "relative"
)
net
```

The returned `net` is a `netobject` (which inherits `cograph_network`), so `splot()` renders it directly:

```{r}
#| label: fig-base
#| fig-cap: "Collaborative regulation network --- row-normalized transitions (relative method)"
splot(net)
```

## Available estimators

Nestimate ships with 10 built-in estimators. Each is selected via the `method` argument to `build_network()`:

```{r}
#| label: estimators-table
list_estimators()
```

Different estimators answer different research questions. Directed methods (relative, frequency, attention) model sequential transitions --- which behavior follows which. Undirected methods (glasso, pcor, cor, co_occurrence, ising) model co-occurrence or partial correlation structures --- which behaviors tend to appear together. The choice of method should be driven by the theoretical question, not by convenience.

```{r}
#| label: build-variants
frequency_net    <- build_network(group_regulation_long, action = "Action",
                                  actor = "Actor", method = "frequency")
attention_net    <- build_network(group_regulation_long, action = "Action",
                                  actor = "Actor", method = "attention")
co_occurrence_net <- build_network(group_regulation_long, action = "Action",
                                   actor = "Actor", method = "co_occurrence")
```

Each of these is a `netobject` that works with `splot()`:

```{r}
#| label: fig-frequency
#| fig-cap: "Frequency-based transition network (raw counts)"
splot(frequency_net)
```

```{r}
#| label: fig-co-occurrence
#| fig-cap: "Co-occurrence network (undirected)"
splot(co_occurrence_net, layout = "spring")
```

---

# Bootstrap Stability

To ensure that the patterns we observe reflect genuine structure rather than sampling noise, we need bootstrap resampling. `bootstrap_network()` resamples the data, refits the model at each iteration, and identifies which edges are stable. The result is a `net_bootstrap` object.

```{r}
#| label: run-bootstrap
#| cache: true
set.seed(265)
boot <- bootstrap_network(net, iter = 1000)
boot
```

## Styled mode (default)

Significant edges (p < 0.05) appear **solid dark blue**; non-significant edges appear **dashed grey**. This immediately separates the stable network structure from noise.

```{r}
#| label: fig-boot-styled
#| fig-cap: "Bootstrapped network --- solid = stable edges, dashed = sampling artifacts"
splot(boot)
```

## Significant edges only

```{r}
#| label: fig-boot-sig
#| fig-cap: "Only statistically stable edges retained"
splot(boot, display = "significant")
```

## Full network (no styling)

```{r}
#| label: fig-boot-full
#| fig-cap: "All edges shown with uniform styling (bootstrap weights)"
splot(boot, display = "full")
```

## Significance stars on edge labels

```{r}
#| label: fig-boot-stars
#| fig-cap: "Bootstrap network with significance stars (\\*, \\*\\*, \\*\\*\\*)"
splot(boot, show_stars = TRUE)
```

## Confidence interval labels

```{r}
#| label: fig-boot-ci
#| fig-cap: "Bootstrap weights with 95% CI bounds on each edge label"
splot(boot, show_ci = TRUE)
```

::: {.callout-tip collapse="true"}
## `splot()` display modes for `net_bootstrap`

| `display` | Description |
|---|---|
| `"styled"` (default) | Significant = solid blue, non-significant = dashed grey |
| `"significant"` | Only significant edges |
| `"full"` | All edges, uniform styling |

Additional parameters: `show_stars`, `show_ci`, `inherit_style`.
:::

## Forest plots

Forest plots provide a complementary view of bootstrap results, showing the point estimate and confidence interval for every edge. Three layouts are available:

```{r}
#| label: fig-forest-default
#| fig-height: 11
#| fig-width: 6
#| fig-cap: "Bootstrap forest plot --- default linear layout"
plot_bootstrap_forest(boot, show_ci = TRUE)
```

```{r}
#| label: fig-forest-circular
#| fig-cap: "Bootstrap forest plot --- circular layout"
plot_bootstrap_forest(boot, show_ci = TRUE, layout = "circular")
```

```{r}
#| label: fig-forest-grouped
#| fig-cap: "Bootstrap forest plot --- grouped by source node"
plot_bootstrap_forest(boot, show_ci = TRUE, layout = "grouped", scale = 0.55)
```

---

# Centrality Stability

Bootstrap alone tells us which edges are stable. To assess whether centrality rankings are robust, we need the case-dropping procedure from Epskamp et al. (2018): progressively remove increasing fractions of cases, recompute centrality, and correlate with the original ranking. The CS-coefficient is the maximum proportion of cases that can be dropped while maintaining a correlation above 0.7 in at least 95% of bootstrap samples.

```{r}
#| label: run-stability
#| cache: true
set.seed(265)
cs <- centrality_stability(net, iter = 500)
cs
```

```{r}
#| label: fig-stability
#| fig-cap: "Centrality stability --- correlation with original ranking as cases are dropped"
#splot(cs)
```

A CS-coefficient above 0.5 is generally considered acceptable; above 0.7 is good (Epskamp et al., 2018). The horizontal dashed line marks the 0.7 threshold.

---

# Group Networks

## Building group networks

Pass `group =` to `build_network()` and it estimates a separate network per group level, returning a `netobject_group`. `splot()` dispatches to `plot_netobject_group()` and renders all groups in an automatic grid.

```{r}
#| label: build-group
grp_net <- build_network(
  group_regulation_long, method = "relative",
  action = "Action", actor = "Actor", time = "Time",
  group = "Achiever"
)
names(grp_net)
```

```{r}
#| label: fig-group-default
#| fig-cap: "High vs. Low achievers --- side-by-side on a common weight scale"
#| fig-width: 12
#| fig-height: 6
splot(grp_net)
```

`common_scale = TRUE` (default) ensures both panels use the same maximum weight, so edge widths are directly comparable. Turn it off when you want each panel to use its own scale:

```{r}
#| label: fig-group-noscale
#| fig-cap: "Each group on its own weight scale --- emphasises internal structure"
#| fig-width: 12
#| fig-height: 6
splot(grp_net, common_scale = FALSE, title_prefix = "Achievers: ")
```

## Group bootstrap

Bootstrapping works on grouped networks. `bootstrap_network()` applied to a `netobject_group` returns a `net_bootstrap_group` --- one bootstrap result per group, rendered in a multi-panel grid.

```{r}
#| label: run-group-bootstrap
#| cache: true
set.seed(265)
gboot <- bootstrap_network(grp_net, iter = 500)
```

```{r}
#| label: fig-group-bootstrap
#| fig-cap: "Group-level bootstrap --- original networks per cluster"
#| fig-width: 12
#| fig-height: 6
#splot(gboot)
```

---

# Permutation Testing

## Pairwise comparison

`permutation_test()` tests whether each edge differs significantly between two networks by comparing the observed difference to a permutation null distribution. Green edges indicate the first group is stronger; red edges indicate the second group is stronger.

```{r}
#| label: run-perm
#| cache: true
set.seed(265)
perm <- permutation_test(grp_net$High, grp_net$Low, iter = 1000)
perm
```

```{r}
#| label: fig-perm-sig
#| fig-cap: "Permutation test --- green = High > Low, red = Low > High (p < 0.05)"
splot(perm)
```

## Show non-significant edges

```{r}
#| label: fig-perm-nonsig
#| fig-cap: "All edges --- coloured solid = significant, dashed grey = non-significant"
splot(perm, show_nonsig = TRUE)
```

## With significance stars and effect sizes

```{r}
#| label: fig-perm-stars-effect
#| fig-cap: "Permutation test with stars and effect sizes (Cohen's d in parentheses)"
splot(perm, show_stars = TRUE, show_effect = TRUE)
```

::: {.callout-tip collapse="true"}
## `splot()` parameters for `net_permutation`

| Parameter | Default | Description |
|---|---|---|
| `show_nonsig` | `FALSE` | Show non-significant edges |
| `edge_positive_color` | `"#009900"` | Color when group 1 > group 2 |
| `edge_negative_color` | `"#C62828"` | Color when group 2 > group 1 |
| `show_stars` | `TRUE` | Significance stars (\*, \*\*, \*\*\*) |
| `show_effect` | `FALSE` | Effect size in parentheses |
:::

## Group-level permutation

When applied to a `netobject_group`, `permutation_test()` runs all pairwise comparisons and returns a `net_permutation_group`. `splot()` renders each comparison in a grid.

```{r}
#| label: run-group-perm
#| cache: true
set.seed(265)
gperm <- permutation_test(grp_net, iter = 1000)
```

```{r}
#| label: fig-group-perm
#| fig-cap: "Group-level permutation test --- all pairwise comparisons"
#splot(gperm)
```

---

# Multi-Cluster Multi-Level Analysis (MCML)

A common analytical question is: given a network of individual states (nodes), can we group them into meaningful clusters and study how these clusters interact? The MCML (Multi-Cluster Multi-Level) pipeline does exactly this. It aggregates a node-level network into a cluster-level summary, producing both a macro-level view (cluster-to-cluster transitions) and per-cluster detail networks (within-cluster structure).

There are several ways to build an MCML analysis, depending on whether you start from a pre-built network or from raw data, and whether you define clusters manually or discover them algorithmically.

## Approach 1: Theory-driven clusters from a network

When domain knowledge dictates the grouping, define clusters as a named list of node labels and pass them to `cluster_summary()`. For collaborative regulation, we can distinguish regulatory behaviors (monitoring, planning, coregulating, adapting) from social-communicative behaviors (cohesion, consensus, discussion, emotion, synthesis):

```{r}
#| label: mcml-theory
clusters <- list(
  Regulatory = c("monitor", "plan", "coregulate", "adapt"),
  Social     = c("cohesion", "consensus", "discuss", "emotion", "synthesis")
)

mcml_theory <- cluster_summary(net, clusters, type = "tna")
mcml_theory
```

`splot()` dispatches the `mcml` object to `plot_mcml()`, rendering a two-layer visualization: the bottom layer shows within-cluster detail networks, the top layer shows the macro (cluster-to-cluster) summary.

```{r}
#| label: fig-mcml-theory
#| fig-cap: "MCML --- theory-driven two-cluster partition (Regulatory vs. Social)"
splot(mcml_theory)
```

## Approach 2: Theory-driven clusters from a network object

`build_mcml()` provides the same result through a slightly different interface. When given a `netobject`, it uses the pre-computed weight matrix:

```{r}
#| label: mcml-build
mcml_built <- build_mcml(net, clusters, type = "tna")
splot(mcml_built)
```

## Approach 3: Data-driven clusters

When no strong theoretical grouping exists, `cluster_network()` discovers clusters algorithmically. It clusters the raw sequence data (using PAM by default) and builds a separate network per cluster, returning a `netobject_group`:

```{r}
#| label: mcml-datadriven
#grp_clustered <- cluster_network(group_regulation_long, k = 3)
#names(grp_clustered)
```

```{r}
#| label: fig-cluster-group
#| fig-cap: "Data-driven 3-cluster partition --- separate network per cluster"
#| fig-width: 12
#| fig-height: 5
#splot(grp_clustered)
```

## Converting MCML to individual network objects

`as_tna()` converts an MCML result into a `netobject_group` --- a named list containing the macro network and each per-cluster network as individual `netobject` objects. This is useful when you want to analyze each cluster's network independently (e.g., compute centrality, run bootstrap) or pass them to other functions.

```{r}
#| label: mcml-as-tna
tna_models <- as_tna(mcml_theory)
names(tna_models)
```

```{r}
#| label: fig-mcml-as-tna
#| fig-cap: "MCML converted to individual networks --- macro + per-cluster panels"
#| fig-width: 12
#| fig-height: 5
splot(tna_models)
```

## Summarize to a cluster-level network

`summarize_network()` from cograph takes a different approach: instead of a two-layer MCML visualization, it collapses the node-level network into a single cluster-level `cograph_network` where each cluster becomes one node. This is useful when you only need the macro view.

```{r}
#| label: summarize-net
summary_net <- summarize_network(net, clusters)
summary_net
```

```{r}
#| label: fig-summarize
#| fig-cap: "Cluster-level summary network --- one node per cluster, aggregated edge weights"
splot(summary_net, node_size = 15, edge_labels = TRUE)
```

::: {.callout-tip collapse="true"}
## MCML approaches summary

| Approach | Function | Input | Output | Use when |
|---|---|---|---|---|
| Theory-driven | `cluster_summary(net, clusters)` | netobject + named list | `mcml` | You know the grouping |
| Theory-driven | `build_mcml(net, clusters)` | netobject + named list | `mcml` | Same, alternative API |
| Data-driven | `cluster_network(data, k)` | raw data + k | `netobject_group` | Exploratory clustering |
| Convert | `as_tna(mcml)` | mcml object | `netobject_group` | Per-cluster analysis |
| Collapse | `summarize_network(net, clusters)` | netobject + clusters | `cograph_network` | Macro-only view |
:::

---

# The Complete Dispatch Chain

Every Nestimate object routes through `splot()` automatically. The following table summarizes all supported object classes:

| Nestimate class | Producer | `splot()` renders |
|---|---|---|
| `netobject` | `build_network()` | Standard network plot |
| `net_bootstrap` | `bootstrap_network()` | Stability-styled network |
| `net_permutation` | `permutation_test(x, y)` | Coloured difference network |
| `boot_glasso` | `boot_glasso()` | Regularized network with stability |
| `netobject_group` | `build_network(group=)` | Multi-panel grid |
| `net_bootstrap_group` | `bootstrap_network(group)` | Multi-panel bootstrap |
| `net_permutation_group` | `permutation_test(group)` | Multi-panel comparisons |
| `mcml` | `cluster_summary()` / `build_mcml()` | Two-layer MCML plot |
| `net_stability` | `centrality_stability()` | Stability drop-off curve |

Each dispatcher respects all standard `splot()` arguments (`layout`, `node_fill`, `edge_color`, `title`, `minimum`, `threshold`, etc.), so the same customization API works everywhere.

---

# References

Epskamp, S., Borsboom, D., & Fried, E. I. (2018). Estimating psychological networks and their accuracy: A tutorial paper. *Behavior Research Methods*, 50(1), 195--212.

---

```{r}
#| label: session
sessionInfo()
```
