---
title: "Higher-Order Network Analysis with Simplicial Complexes"
subtitle: "From Transition Networks to Topological Analysis"
author: "Mohammed Saqr & Sonsoles López-Pernas"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    theme: cosmo
    fig-width: 10
    fig-height: 7
execute:
  warning: false
  message: false
---

## Introduction

When we analyze sequential data as a network --- who follows whom, which state leads to which --- we typically build a **first-order Markov model**: the probability of the next state depends only on the current state. This is the foundation of Transition Network Analysis (TNA).

But human behavior is not always memoryless. Consider two programmers working with an AI coding assistant:

- Programmer A: *Specify → Request → Specify → Request* (a steady back-and-forth)
- Programmer B: *Frustrate → Specify → Request → Command* (escalating from frustration to direct commands)

In a first-order model, both sequences contribute equally to the *Specify → Request* transition. But the **context** --- what came before --- matters. The sequence *Frustrate → Specify* may lead to very different next actions than *Request → Specify*. These are **higher-order dependencies**: cases where the past two (or more) states jointly predict the future better than the current state alone.

This tutorial introduces a complete pipeline for detecting, quantifying, and visualizing higher-order structure in sequential data. We progress from standard transition networks through statistical order detection, pathway analysis, and into the topology of higher-order interactions using simplicial complexes.

### What you will learn

1. Test whether your data has higher-order dependencies (and at what order)
2. Identify which specific pathways deviate from first-order expectations
3. Detect anomalously frequent or rare behavioral sequences
4. Compute topological invariants (Betti numbers, Euler characteristic)
5. Track how network topology changes across thresholds (persistent homology)
6. Visualize higher-order pathways as simplicial blob diagrams
7. Compare higher-order structure across groups

### Packages

```{r setup}
library(Nestimate)
library(cograph)
```

**cograph** provides the simplicial visualization (`plot_simplicial()`). **Nestimate** provides the higher-order network construction (`build_hon`, `build_hypa`, `build_mogen`) and simplicial complex analysis (`build_simplicial`, `betti_numbers`, `persistent_homology`, `q_analysis`).

## Building the Network

We use a dataset of human--AI pair programming sessions where a researcher coded each human action into 9 behavioral categories: Request, Specify, Command, Correct, Refine, Verify, Inquire, Interrupt, and Frustrate.

```{r build-network}
data("human_cat", package = "Nestimate")

net <- build_network(human_cat, method = "relative",
                     action = "category", actor = "session_id",
                     time = "timestamp")
net
```

```{r plot-network, fig.width=9, fig.height=9}
splot(net, minimum = 0.05, title = "Human Actions in AI-Assisted Coding")
```

Several patterns stand out: **Specify** is the hub --- most actions lead to and from it. Self-loops on Specify and Command indicate persistence. The **Request → Specify → Command** backbone reflects the natural workflow. But this first-order view might be hiding important patterns. Does it matter *how* someone arrived at Specify?

## Is First-Order Enough? Multi-Order Model Selection

Before investing in higher-order analysis, we should answer: **does sequential memory actually matter?** The Multi-Order Generative Model (MOGen; Scholtes 2017) fits Markov models at increasing orders and compares them using information criteria. Higher-order models fit better but have exponentially more parameters --- AIC and BIC find the best tradeoff.

```{r mogen}
mg <- build_mogen(net, max_order = 4)
summary(mg)
```

```{r plot-mogen, fig.width=8, fig.height=5}
plot(mg)
```

### Interpreting the results

- **Order 0**: Actions are independent --- no sequential structure.
- **Order 1**: The standard Markov model is sufficient.
- **Order 2+**: Higher-order dependencies exist. A first-order TNA would miss them.

Even when order 1 is globally optimal (by BIC), *specific pathways* may still exhibit higher-order behavior --- the global model averages over all states. The HON analysis below detects these local patterns.

### Second-order transition matrix

`mogen_transitions()` extracts the transitions at a given order, showing which two-step contexts lead to different outcomes:

```{r mogen-transitions}
mt <- mogen_transitions(mg, order = 2)
mt[1:8, ]
```

Each row is a second-order pathway: the `from` column shows the two-state context, `to` is the predicted next state, and `probability` is the conditional probability given that context.

## Higher-Order Networks (HON)

While MOGen tells us *whether* higher-order dependencies exist on average, the **Higher-Order Network** (HON; Xu et al. 2016) tells us *where*. HON tests, for each state, whether adding context significantly changes the outgoing transition distribution using KL-divergence. If the distribution after *Request → Specify* differs from *Frustrate → Specify*, HON creates separate nodes for these contexts.

```{r hon}
hon <- build_hon(net)
hon
```

```{r hon-detail}
ho_edges <- hon$edges[hon$edges$from_order > 1, ]
ho_top <- ho_edges[order(-ho_edges$count), ]
ho_top[1:10, ]
```

Out of `r nrow(hon$edges)` total edges, **`r nrow(ho_edges)`** (`r round(100 * nrow(ho_edges) / nrow(hon$edges))`%) are higher-order --- transitions where the preceding context changes the outcome. Look for pathways where the higher-order probability differs substantially from the first-order probability --- these represent genuine contextual effects.

### Raw path frequencies

The most common 3-step sequences regardless of statistical significance:

```{r paths}
path_counts(net, k = 3, top = 15)
```

## Path Anomaly Detection (HYPA)

HON detects *where* higher-order structure exists. **HYPA** (LaRock et al. 2020) takes a complementary approach: it detects which specific paths occur *significantly more or less often than expected* under a null model.

HYPA constructs a De Bruijn graph and computes expected path frequencies using a multi-hypergeometric null model. Paths that deviate significantly are flagged:

- **Over-represented** paths: These sequences occur far more often than chance predicts. They represent **learned behavioral routines**.
- **Under-represented** paths: Systematically avoided combinations --- sequences that are socially inappropriate, cognitively difficult, or practically ineffective.

```{r hypa}
#| fig-height: 9
#| fig-width: 15
hypa <- build_hypa(net)


plot_simplicial(hypa, dismantled = TRUE)

```

### Over-represented pathways

Learned routines --- sequences that occur far more than expected. `plot_simplicial()` accepts HYPA objects directly with the `anomaly` parameter to filter:

```{r hypa-over, fig.width=14, fig.height=14}
plot_simplicial(hypa, anomaly = "over", max_pathways = 9,
                dismantled = TRUE, ncol = 3)
```

### Under-represented pathways

Avoided sequences:

```{r hypa-under, fig.width=14, fig.height=7}
plot_simplicial(hypa, anomaly = "under", max_pathways = 6,
                dismantled = TRUE, ncol = 3)
```

### Combined overlay

Multiple pathways on a single plot --- overlapping regions indicate states that participate in many higher-order patterns:

```{r simplicial-combined, fig.width=10, fig.height=10}
plot_simplicial(net, max_pathways = 6,
                title = "Top 6 Higher-Order Pathways")
```

## Higher-Order Analysis Across Groups

A powerful application is comparing higher-order structure across conditions. We build grouped networks by superclass (Directive, Evaluative, Metacognitive) and run HYPA on each:

```{r group-hypa}
grp_net <- build_network(human_cat, method = "relative",
                         action = "category", actor = "session_id",
                         time = "timestamp", group = "superclass")

hypa_results <- lapply(names(grp_net), function(nm) {
  h <- build_hypa(grp_net[[nm]])
  data.frame(
    group = nm,
    total_edges = h$n_edges,
    anomalous = h$n_anomalous,
    over = h$n_over,
    under = h$n_under,
    pct_anomalous = round(100 * h$n_anomalous / h$n_edges, 1)
  )
})
do.call(rbind, hypa_results)
```

Groups with a higher percentage of anomalous pathways have more higher-order structure --- their sequential dynamics are less predictable from first-order transitions alone.

```{r group-simplicial, fig.width=14, fig.height=7}
par(mfrow = c(1, 2))
plot_simplicial(grp_net$Directive, max_pathways = 4,
                title = "Directive: Higher-Order Pathways")
plot_simplicial(grp_net$Evaluative, max_pathways = 4,
                title = "Evaluative: Higher-Order Pathways")
```

## Comparing Datasets

Higher-order analysis is particularly informative when applied to different datasets. Here we compare the human-AI coding data with a collaborative regulation dataset:

```{r compare-datasets}
data("group_regulation_long", package = "Nestimate")
net_reg <- build_network(group_regulation_long, method = "relative",
                         action = "Action", actor = "Actor")

hypa_coding <- build_hypa(net)
hypa_reg <- build_hypa(net_reg)

comparison <- data.frame(
  dataset = c("Human-AI Coding", "Collaborative Regulation"),
  nodes = c(net$n_nodes, net_reg$n_nodes),
  total_paths = c(hypa_coding$n_edges, hypa_reg$n_edges),
  anomalous = c(hypa_coding$n_anomalous, hypa_reg$n_anomalous),
  over = c(hypa_coding$n_over, hypa_reg$n_over),
  under = c(hypa_coding$n_under, hypa_reg$n_under)
)
comparison
```

```{r compare-simplicial, fig.width=14, fig.height=7}
par(mfrow = c(1, 2))
plot_simplicial(net, max_pathways = 4,
                title = "Human-AI Coding")
plot_simplicial(net_reg, max_pathways = 4,
                title = "Collaborative Regulation")
```

Different domains produce different higher-order signatures. Comparing these reveals whether sequential memory is a universal property of the behavioral system or specific to particular tasks.

## Simplicial Complex Analysis

All the methods above analyze sequential pathways. **Simplicial complex analysis** takes a different perspective: it examines the **topological structure** of the network itself.

### What is a simplicial complex?

A standard network captures pairwise relationships: node A connects to node B. But many interactions involve **groups**: states A, B, and C may all be densely interconnected, forming a triangle. A simplicial complex formalizes this:

- A **0-simplex** is a single node
- A **1-simplex** is an edge (two connected nodes)
- A **2-simplex** is a filled triangle (three mutually connected nodes)
- A **$k$-simplex** is a group of $k+1$ mutually connected nodes

The **clique complex** turns every clique into a simplex, lifting a flat graph into a multi-dimensional topological object.

### Building the complex

```{r simplicial-build}
sc <- build_simplicial(net, threshold = 0.05)
sc
```

### Topological descriptors

- **f-vector** $(f_0, f_1, f_2, \ldots)$: counts simplices at each dimension.
- **Betti numbers** $(\beta_0, \beta_1, \beta_2, \ldots)$: the central invariants. $\beta_0$ = connected components, $\beta_1$ = independent loops, $\beta_2$ = enclosed voids.
- **Euler characteristic** $\chi = \sum (-1)^k f_k = \sum (-1)^k \beta_k$. For a contractible complex, $\chi = 1$.
- **Simplicial degree**: how many higher-dimensional simplices each node participates in.

```{r simplicial-degree}
simplicial_degree(sc)
```

```{r plot-sc, fig.width=13, fig.height=9}
plot(sc)
```

Nodes with high simplicial degree are topological hubs --- they participate in many group interactions. This captures group-level importance that pairwise degree misses.

### Persistent Homology

A single complex at one threshold gives a static snapshot. **Persistent homology** reveals how topology evolves as we vary the threshold. Start with only the strongest edges and progressively add weaker ones. Features that persist across a wide range are **topologically robust**.

```{r persistent, fig.width=13, fig.height=6}
ph <- persistent_homology(net, n_steps = 25)
ph
plot(ph)
```

**Betti curve** (left): At high threshold, the network is sparse (many components). As threshold decreases, components merge ($\beta_0$ drops) and loops may form ($\beta_1$ rises) then get filled.

**Persistence diagram** (right): Points far from the diagonal have long lifetimes --- structurally important features. Points near the diagonal are noise.

### Q-Analysis

Q-analysis (Atkin 1974) measures connectivity at multiple levels of structural sharing. Two maximal simplices are **q-connected** if they share a face of dimension $\geq q$:

- **q = 0**: share at least one node
- **q = 1**: share at least one edge
- **q = 2**: share at least one triangle

```{r q-analysis, fig.width=13, fig.height=6}
qa <- q_analysis(sc)
qa
plot(qa)
```

The **fragmentation point** (first q where components > 1) is critical. Below it, the network is fully connected through shared sub-structures. Above it, parts become structurally isolated.

### Pathway Complex from HON

An alternative to the clique complex: build a simplicial complex from the **pathways** discovered by HON --- topology derived from sequential dynamics rather than network structure:

```{r pathway-complex, fig.width=13, fig.height=9}
sc_hon <- build_simplicial(hon, type = "pathway", max_pathways = 30)
sc_hon
plot(sc_hon)
```

In the pathway complex, each higher-order pathway becomes a simplex. $\beta_0 > 1$ means some states are sequentially disconnected. $\beta_1 > 0$ means there are cyclic pathway structures.

### Verification

All simplicial computations are cross-validated against igraph's clique-finding and verified via the Euler-Poincaré theorem:

```{r verify}
verify_simplicial(net$weights, threshold = 0.05)
```

## Summary

| Step | Method | Function | What it reveals |
|------|--------|----------|-----------------|
| 1 | **TNA** | `build_network()` | First-order transition structure |
| 2 | **MOGen** | `build_mogen()` | Whether higher-order is needed |
| 3 | **HON** | `build_hon()` | Where sequential context changes transitions |
| 4 | **HYPA** | `build_hypa()` | Which paths are anomalously frequent or rare |
| 5 | **Visualization** | `plot_simplicial()` | Blob diagrams of pathways |
| 6 | **Simplicial** | `build_simplicial()` | Topological structure |
| 7 | **Persistence** | `persistent_homology()` | Robustness across scales |
| 8 | **Q-analysis** | `q_analysis()` | Multi-level connectivity |

The key progression:

- **TNA** answers: *what transitions exist?*
- **MOGen** answers: *is first-order enough?*
- **HON** answers: *where does context matter?*
- **HYPA** answers: *which sequences are surprising?*
- **Simplicial analysis** answers: *what is the shape of the interaction space?*

## References

- Atkin, R. H. (1974). *Mathematical Structure in Human Affairs*. Heinemann Educational.
- Battiston, F., et al. (2020). Networks beyond pairwise interactions: Structure and dynamics. *Physics Reports*, 874, 1--92.
- LaRock, T., Scholtes, I., & Eliassi-Rad, T. (2020). HYPA: Efficient detection of path anomalies in time series data on networks. *EPJ Data Science*, 9(1), 15.
- Scholtes, I. (2017). When is a network a network? Multi-order graphical model selection in pathways and temporal networks. *Proceedings of the 23rd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining*, 1037--1046.
- Xu, J., Wickramarathne, T. L., & Chawla, N. V. (2016). Representing higher-order dependencies in networks. *Science Advances*, 2(5), e1600028.
