---
title: "Comparing two networks with compare_networks()"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6)
library(Nestimate)
```

High and low achievers in `group_regulation_long` are compared with a
permutation test (1,000 permutations).

```{r}
achievers <- build_network(group_regulation_long, method = "relative",
                           actor = "Actor", action = "Action",
                           time = "Time", group = "Achiever")
cmp <- compare_networks(achievers, test = "permutation", iter = 1000, seed = 1)
cmp
summary(cmp)
```

The networks are similar overall (Pearson 0.92, cosine 0.95, Jaccard
0.75) but differ in 42 of 78 transitions and 16 of 27 node measures. The
global test is significant (M = 2.61, p = 0.001).

## networks
```{r}
plot(cmp, type = "networks")
```

The two networks side by side.

## difference
```{r}
plot(cmp, type = "difference")
```

High minus Low for each transition: blue is higher in High, red higher in
Low; the line style also carries the sign.

## edges
```{r}
plot(cmp, type = "edges")
```

The largest differences: `discuss` to `consensus` (0.42 vs 0.21) and
`synthesis` to `consensus` (0.58 vs 0.39) are higher in High; `synthesis`
to `adapt` is higher in Low (0.14 vs 0.30). All have p = 0.001.

## nodes
```{r}
plot(cmp, type = "nodes")
```

Centrality differences per state. `synthesis` has higher betweenness in
High (7 vs 0, p = 0.003); `adapt` in Low (4 vs 1, p = 0.001).

## global
```{r}
plot(cmp, type = "global")
```

The 22 similarity and distance metrics of `global_differences(cmp)`,
grouped by category.

## heatmap
```{r}
plot(cmp, type = "heatmap")
```

Every cell of the High minus Low difference matrix, with its value.

## scatter
```{r}
plot(cmp, type = "scatter")
```

Each transition's weight in High against Low; points on the diagonal
are equal in both (Pearson 0.92).

## inference
```{r}
plot(cmp, type = "inference")
```

Edge differences with their permutation p-values; filled markers are
significant at 0.05.

## Accounting for nesting
```{r}
nested <- compare_networks(achievers, test = "permutation", iter = 1000,
                           seed = 1, actor = "Group")
nested
tail(global_differences(nested, digits = 3), 5)
```

Students are nested in teams, and the achievement level is given per
team. With `actor = "Group"`, the permutation test reassigns whole teams
instead of single students, and `global_differences()` gains three
`Nesting` rows. The ICC is −0.002, which indicates little evidence of a
nesting effect; the design effect is 1.01 for the transitions and 1.17
for M. The networks still differ overall (M = 2.61, p = 0.001), and 42
transitions remain significant.
