---
title: "Nested sequences and the permutation test"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(Nestimate)
```

Do high and low achievers regulate their group work differently, or do
their teams? In `group_regulation_long`, 2,000 students work in 200 teams
of 10, and the achievement level is given to the team, not to the
student. A comparison of High and Low is therefore a comparison of teams.
The students are nested in teams. The permutation test treats each
student's sequence as an independent unit, so a nesting effect, teammates
behaving alike, would enter the comparison. This tutorial runs the
comparison, checks for a nesting effect, and shows how to report it.

```{r data}
achievers <- build_network(group_regulation_long, method = "relative",
                           actor = "Actor", action = "Action",
                           time = "Time", group = "Achiever")
achievers
```

`build_network()` builds one transition network per achievement level
from the students' action sequences. Both networks have the same nine
regulation actions and 75 to 76 non-zero transitions.

## The ordinary test

```{r plain}
permutation(achievers, iter = 1000, seed = 1)
```

`permutation()` pools the 2,000 sequences, reassigns them to High and Low
1,000 times, and compares the observed differences with the reassigned
ones. Two kinds of result are reported. The edge tests ask, for each
transition, whether the High minus Low difference is larger than the
reassignments produce: 42 of 78 transitions are. The global test asks
whether the two networks differ at all, with two statistics from the
Network Comparison Test (van Borkulo et al., 2023): M, the sum of all
absolute edge differences, and S, the largest one. Both have
p = 0.000999, the smallest value 1,000 reassignments can give: no
reassignment produced a difference as large as the observed one.

The test assumes that any student could have been in either group. In
this design only whole teams could, so the next step accounts for the
nesting.

## Reassigning whole teams

```{r actor}
permutation(achievers, iter = 1000, seed = 1, actor = "Group")
```

`actor = "Group"` tells the test that the sequences are nested in teams.
Each reassignment now moves all ten students of a team together, which is
how the achievement levels were actually given. The observed differences
are the same; only the reference distribution changes.

The edge tests and the global test give the same result as with
student-level reassignment: 42 significant transitions, p = 0.000999 for
M and S. The output also reports the intraclass correlation,
ICC = −0.002 (95% CI −0.006 to 0.002), and the design effect, 1.03 for
the transitions and 1.20 for M. Both are defined and interpreted in the
next section.

A low ICC means that sequences from the same team are about as different
from each other as sequences from different teams: the observations are
close to independent, and an analysis that ignores the nesting gives
nearly the same results as one that accounts for it. Accounting for
nesting becomes necessary when the ICC is not negligible, and more so
when units are large, because the design effect grows with both:
approximately 1 + (m − 1) × ICC for units of size m (Kish, 1965). A
commonly cited rule of thumb is that design effects below 2 do not lead
to seriously misleading inferences (Muthén & Satorra, 1995). Here the
ICC is −0.002 and the design effect 1.03, so accounting for the nesting
was not necessary, and the results with and without it were almost the
same.

## The nesting diagnostics

The intraclass correlation (ICC) is the proportion of the total variance
that lies between units rather than within them (Shrout & Fleiss, 1979):

$$\text{ICC} = \frac{\sigma^2_{\text{between}}}{\sigma^2_{\text{between}} + \sigma^2_{\text{within}}}.$$

An ICC of 0 means that members of the same unit are no more alike than
members of different units; an ICC of 1 means that all members of a unit
are identical and all variation lies between units. A value in between is
the proportion of variation due to the unit. In `permutation()` the ICC is
computed for each transition: for every student, the share of their
transitions that is, for example, `plan` to `monitor`; the one-way ANOVA
ICC of that share across teams is computed within High and within Low.
The transition ICCs are averaged with weights equal to transition
frequency. The 95% interval comes from the jackknife, leaving out one
team at a time (Efron & Tibshirani, 1993).

The design effect is the ratio of the variance of an estimate under the
nested design to its variance had the units been sampled independently
(Kish, 1965). Here it is the variance of the reassigned differences when
whole teams move, divided by the variance when single students move. A
value of 1 means both reassignments give the same chance variation.

`permutation_diagnostics()` runs the ordinary and the team-level test on
the same data and reports both with these two quantities, one row per
compared pair:

```{r diagnostics}
permutation_diagnostics(achievers, actor = "Group", iter = 1000, seed = 1)
```

The comparison involves 2,000 sequences in 200 teams. The design is
`between`: every team belongs to one achievement level (`within` would mean
units in both groups, as in repeated measures; `mixed`, some of each). The
ICC is −0.002 with a 95% interval from −0.006 to 0.002, which indicates
little evidence of a nesting effect; a value slightly below zero is
sampling variation around zero. The design effect is 1.03 for the
transitions (median over edges) and 1.20 for the global statistic M. The
global test has p = 0.000999 under both reassignments, and 42 transitions
are significant under both. Two transitions change their conclusion. The
smallest p-value the team-level test can produce, `min_p_actor`, is
0.000999: with 200 teams the number of iterations sets the limit, not the
number of teams.

With `level = "edges"` the same comparison is reported per transition:

```{r edges}
head(permutation_diagnostics(achievers, actor = "Group", iter = 1000,
                             level = "edges", seed = 1), 6)
```

`diff` is the High minus Low transition probability and is the same under
both tests. `icc` is the ICC of that transition. `null_sd_sequence` and
`null_sd_actor` are the spread of the reassigned differences under
student-level and team-level reassignment, and `deff` is the squared ratio
of the two. `p_sequence` and `p_actor` are the two edge p-values, and
`changed` marks a transition that is significant at 0.05 under one test
but not the other. In the six rows shown, the ICCs lie between −0.009 and
0.010 and the design effects between 0.82 and 1.14; the two spreads are
close, so the two p-values are close. `adapt` to `coregulate` has
p = 0.015 and p = 0.024 and is significant under both.

Over all 78 transitions, two change at this seed (`coregulate` to `adapt`,
`emotion` to `plan`), with p-values between 0.047 and 0.072 under both
tests. With seeds 2 to 4 a different set of two or three transitions
changes, all with p-values between 0.04 and 0.07. These changes come from
the finite number of reassignments, not from the nesting. The ICC
indicates little evidence of a nesting effect, and the High–Low
difference does not depend on the nesting of students in teams.

## Reporting

> Because the data were nested (students in teams; with repeated
> measurements, sessions in students), the two groups were compared with
> a permutation test that accounted for the nesting: in each of 1,000
> permutations, whole teams rather than single sequences were reassigned
> between the groups, so that sequences from the same team stayed
> together (Good, 2005). Units that belonged to one group were reassigned
> as a whole, and units that contributed to both groups, as in a
> repeated-measures design, had their sequences reassigned within the
> unit. The nesting effect was quantified with the intraclass correlation
> (ICC; Shrout & Fleiss, 1979) and the design effect, the ratio of the
> variance under the nested design to the variance under independent
> sampling (Kish, 1965). The ICC indicated little evidence of a nesting
> effect (ICC = −0.002, 95% CI −0.006 to 0.002; design effect 1.03 across
> 200 teams). The networks of high and low achievers differed overall
> (M = 2.61, p < .001), and 42 of 78 transitions differed at p < .05.

## When to use which

- `permutation()`: one sequence per student, with no grouping above the
  student.
- `permutation(actor = )`: several sequences per student, or students in
  teams, classes or other units. It handles units in one group, units in
  both groups, and mixtures, and reports the ICC and design effect.
- `paired = TRUE`: exactly one sequence per student in each of two
  conditions.
- `permutation_diagnostics()`: to show next to each other what the
  ordinary and the team-level test conclude.
- `actor` applies to transition networks (`relative`, `frequency`,
  `co_occurrence`). Correlation networks do not keep the unit
  identifiers after estimation.

## References

Kish, L. (1965). *Survey Sampling*. Wiley.

Muthén, B. O., & Satorra, A. (1995). Complex sample data in structural
equation modeling. *Sociological Methodology*, 25, 267–316.

Shrout, P. E., & Fleiss, J. L. (1979). Intraclass correlations: Uses in
assessing rater reliability. *Psychological Bulletin*, 86(2), 420–428.

van Borkulo, C. D., van Bork, R., Boschloo, L., Kossakowski, J. J.,
Tio, P., Schoevers, R. A., Borsboom, D., & Waldorp, L. J. (2023).
Comparing network structures on three aspects: A permutation test.
*Psychological Methods*, 28(6), 1273–1285.

Efron, B., & Tibshirani, R. J. (1993). *An Introduction to the
Bootstrap*. Chapman & Hall.

Good, P. (2005). *Permutation, Parametric, and Bootstrap Tests of
Hypotheses* (3rd ed.). Springer.
