---
title: "Network Estimation with Nestimate"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
options(max.print = 80)
library(Nestimate)
```

# Purpose

`build_network()` is the main entry point in Nestimate. It accepts raw
sequence data, long event logs, grouped event logs, one-hot activity data, and
numeric data for psychological networks. The return value is usually a
`netobject`, a common S3 structure used by the rest of the package.

This tutorial starts with transition networks, then moves to co-occurrence,
window-based TNA, grouped networks, psychological networks, and validation.

# Event Logs and Transition Networks

The `human_long` data set is a long event log. Each row is one coded action
from a human-AI programming session.

```{r data-human}
data(human_long)
str(human_long)
table(human_long$cluster)
```

The network needs three columns: an action label, an actor or sequence id, and
an ordering variable.

```{r transition-network}
net_relative <- build_network(
  human_long,
  method = "relative",
  action = "code",
  actor = "session_id",
  time = "timestamp"
)

net_relative
round(net_relative$weights[1:5, 1:5], 3)
```

`method = "relative"` estimates row-normalized transition probabilities. Each
row is the current state; each column is the next state. A large value means
that the next state often follows the row state.

The object keeps the prepared sequence data. This matters because downstream
functions such as clustering and bootstrap can reuse it.

```{r netobject-fields}
names(net_relative)
dim(net_relative$data)
head(net_relative$data[, 1:6])
```

# Frequency Networks

Use `method = "frequency"` when the absolute number of transitions is needed.
This is the right input for count-based procedures, including chi-square
mosaic plots.

```{r frequency-network}
net_frequency <- build_network(
  human_long,
  method = "frequency",
  action = "code",
  actor = "session_id",
  time = "timestamp"
)

net_frequency
round(net_frequency$weights[1:5, 1:5], 0)
```

The largest edges in a frequency network are the transitions that happen most
often, not necessarily the transitions with the highest conditional
probability.

```{r top-frequency-edges}
head(net_frequency$edges[order(-net_frequency$edges$weight), ], 10)
```

# Attention-Weighted Transitions

Attention-weighted TNA gives later transitions more influence. This is useful
when the final part of a process is more informative than the beginning.

```{r attention-network}
net_attention <- build_network(
  human_long,
  method = "attention",
  action = "code",
  actor = "session_id",
  time = "timestamp"
)

net_attention
round(net_attention$weights[1:5, 1:5], 3)
```

Read the output as a directed transition network, but remember that transition
weights include temporal decay.

# Co-occurrence Networks

Co-occurrence networks answer a different question: which states tend to occur
together in the same sequence or window?

```{r cooccurrence-network}
net_co <- build_network(
  human_long,
  method = "co_occurrence",
  action = "cluster",
  actor = "session_id",
  time = "timestamp"
)

net_co
round(net_co$weights, 3)
```

This network is undirected. It is not a model of what follows what.

# Window-Based TNA

`wtna()` works with one-hot activity data. The bundled
`learning_activities` data has one row per student-time window and one column
per activity.

```{r wtna-data}
data(learning_activities)
head(learning_activities)
```

```{r wtna-transition}
net_wtna <- wtna(
  learning_activities,
  actor = "student",
  method = "transition",
  type = "relative"
)

net_wtna
round(net_wtna$weights, 3)
```

Use `method = "cooccurrence"` for within-window co-activity and
`method = "both"` when temporal succession and co-activity should be kept
together.

```{r wtna-both}
net_wtna_both <- wtna(
  learning_activities,
  actor = "student",
  method = "both",
  type = "relative"
)

net_wtna_both
```

# Grouped Networks

Passing `group =` estimates one network per group and returns a
`netobject_group`.

```{r grouped-networks}
data(group_regulation_long)

course_net <- build_network(
  group_regulation_long,
  method = "relative",
  action = "Action",
  actor = "Actor",
  time = "Time",
  group = "Course"
)

course_net
names(course_net)
lapply(course_net, function(x) dim(x$weights))
```

This is a useful form when groups are known in advance. For discovered groups,
use clustering first, as shown in the clustering tutorial.

# Psychological Networks

For numeric data, `build_network()` estimates association networks.

```{r psychological-data}
data(chatgpt_srl)
str(chatgpt_srl)
```

```{r psychological-networks}
net_cor <- build_network(chatgpt_srl, method = "cor")
net_pcor <- build_network(chatgpt_srl, method = "pcor")
net_glasso <- build_network(chatgpt_srl, method = "glasso",
                            params = list(gamma = 0.5))

net_glasso
round(net_glasso$weights, 3)
```

The three methods answer progressively stricter questions:

- `cor`: which variables are associated pairwise?
- `pcor`: which variables remain associated after controlling for others?
- `glasso`: which partial correlations survive regularization?

Predictability is available for these undirected association methods.

```{r predictability}
predictability(net_glasso)
```

# Validation

Validation functions operate on the network object and reuse the stored data.

```{r reliability}
set.seed(1)
network_reliability(net_relative, iter = 50)
```

```{r bootstrap}
set.seed(1)
boot <- bootstrap_network(net_relative, iter = 50)
boot
```

```{r centrality-stability}
set.seed(1)
cs <- centrality_stability(net_relative, iter = 50)
cs
```

For grouped networks, use permutation testing to compare edges.

```{r permutation-group}
set.seed(1)
perm <- permutation(course_net, iter = 50)
perm
```

# A Practical Reading Order

Start with the network type that matches the question:

- For temporal order, use `relative`, `frequency`, or `attention`.
- For co-activity, use `co_occurrence` or `wtna(..., method = "cooccurrence")`.
- For psychological variables, use `cor`, `pcor`, `glasso`, `ising`, or `mgm`.
- For known groups, use `group =`.
- For unknown groups, use clustering or MMM before building per-cluster
  networks.

Then validate before interpreting individual edges.

