---
title: "Sequence Clustering in Nestimate"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
options(max.print = 80)
library(Nestimate)
```

# Purpose

Clustering separates sequences into groups with similar behavior. In
Nestimate there are two families:

1. Distance-based clustering with `build_clusters()`.
2. Model-based clustering with `build_mmm()`.

The first compares observed sequences directly. The second fits a mixture of
Markov chains and clusters sequences by transition dynamics.

# Build a Sequence Network First

We use a small sample of `human_long` to keep the rendered tutorial responsive.
The clustering works on the stored sequence data inside the network object.

```{r data}
data(human_long)

set.seed(2)
sessions <- sample(unique(human_long$session_id), 120)
human_sub <- human_long[human_long$session_id %in% sessions, ]

net <- build_network(
  human_sub,
  method = "relative",
  action = "cluster",
  actor = "session_id",
  time = "timestamp"
)

net
dim(net$data)
head(net$data[, 1:8])
```

# Distance-Based Clustering

`build_clusters()` computes a distance matrix between sequences, then applies
PAM or hierarchical clustering.

```{r build-clusters}
cl <- build_clusters(
  net,
  k = 3,
  dissimilarity = "hamming",
  method = "pam",
  seed = 1
)

cl
summary(cl)
```

The main quantities are:

- `assignments`: one cluster label per sequence.
- `sizes`: number of sequences per cluster.
- `silhouette`: overall separation quality.
- `distance`: the pairwise dissimilarity matrix.
- `medoids`: representative sequences when `method = "pam"`.

```{r cluster-fields}
head(cl$assignments, 20)
cl$sizes
cl$silhouette
```

# Inspect the Fit

The silhouette plot shows whether each sequence fits its assigned cluster.

```{r silhouette-plot}
plot(cl, type = "silhouette")
```

The MDS plot projects the distance matrix to two dimensions. It is a visual
summary, not the clustering algorithm itself.

```{r mds-plot}
plot(cl, type = "mds")
```

The heatmap shows the distance matrix ordered by cluster assignment.

```{r heatmap-plot}
plot(cl, type = "heatmap")
```

# Choose the Distance Metric

Different distances encode different assumptions.

```{r compare-distances}
cl_hamming <- build_clusters(net, k = 3, dissimilarity = "hamming",
                             method = "ward.D2")
cl_lcs <- build_clusters(net, k = 3, dissimilarity = "lcs",
                         method = "ward.D2")
cl_lv <- build_clusters(net, k = 3, dissimilarity = "lv",
                        method = "ward.D2")

data.frame(
  distance = c("hamming", "lcs", "lv"),
  silhouette = c(cl_hamming$silhouette, cl_lcs$silhouette, cl_lv$silhouette),
  min_size = c(min(cl_hamming$sizes), min(cl_lcs$sizes), min(cl_lv$sizes)),
  max_size = c(max(cl_hamming$sizes), max(cl_lcs$sizes), max(cl_lv$sizes))
)
```

Use Hamming when position-by-position alignment is meaningful. Use edit
distances when insertions or shifts are meaningful. Use LCS when shared order
matters more than exact timing.

Weighted Hamming gives more influence to earlier positions.

```{r weighted-hamming}
cl_weighted <- build_clusters(
  net,
  k = 3,
  dissimilarity = "hamming",
  weighted = TRUE,
  lambda = 0.5,
  method = "ward.D2"
)

cl_weighted
```

# Sweep Choices With `cluster_choice()`

Manual one-off comparisons are fine for learning. For real work, sweep the
candidate space explicitly.

```{r cluster-choice}
choice <- cluster_choice(
  net,
  k = 2:5,
  dissimilarity = c("hamming", "lcs", "lv"),
  method = c("pam", "ward.D2"),
  seed = 1
)

choice
```

```{r cluster-choice-plot}
plot(choice, type = "facet", abbrev = TRUE)
```

Silhouette is useful, but do not choose from it alone. A high silhouette with
a very small cluster may be a poor substantive solution.

```{r cluster-choice-tradeoff}
plot(choice, type = "tradeoff", abbrev = TRUE)
```

# Diagnostics

`cluster_diagnostics()` gives a common diagnostic surface for distance
clustering and MMM.

```{r diagnostics}
diag <- cluster_diagnostics(cl_hamming)
diag
as.data.frame(diag)
```

```{r diagnostics-plot}
plot(diag, type = "silhouette")
```

# Build Per-Cluster Networks

`build_clusters()` returns clustering results, not per-cluster networks. Pass
the clustering object to `build_network()` to estimate one network per cluster.

```{r build-cluster-networks}
cluster_nets <- build_network(cl_hamming, method = "relative")
cluster_nets
names(cluster_nets)
```

Each member is a normal `netobject`.

```{r cluster-network-mats}
lapply(cluster_nets, function(x) round(x$weights, 2))
```

The one-call wrapper is `cluster_network()`.

```{r cluster-network-wrapper}
cluster_nets2 <- cluster_network(
  net,
  k = 3,
  cluster_by = "ward.D2",
  dissimilarity = "hamming"
)

cluster_nets2
```

# Model-Based Clustering With MMM

Mixed Markov Models cluster sequences by transition dynamics. Each component
has its own transition matrix.

```{r mmm}
set.seed(1)
mmm <- build_mmm(net, k = 2, n_starts = 5, max_iter = 80, seed = 1)
mmm
summary(mmm)
```

Posterior probabilities show assignment certainty.

```{r mmm-posterior}
plot(mmm, type = "posterior")
```

Compare component counts with `compare_mmm()`.

```{r compare-mmm}
cmp <- compare_mmm(net, k = 2:4, n_starts = 3, max_iter = 60, seed = 1)
cmp
plot(cmp)
```

Build per-component networks with `cluster_mmm()`.

```{r cluster-mmm}
mmm_nets <- cluster_mmm(net, k = 2, n_starts = 5, max_iter = 80, seed = 1)
mmm_nets
cluster_diagnostics(mmm_nets)
```

# Covariates

Distance clustering treats covariates as post-hoc descriptors. They do not
change the cluster assignments.

```{r covariates-distance, eval=requireNamespace("nnet", quietly = TRUE)}
cl_cov <- build_clusters(
  net,
  k = 3,
  method = "ward.D2",
  covariates = c("project")
)

summary(cl_cov)
```

MMM covariates are different. They are part of the model and affect the mixing
proportions.

```{r covariates-mmm, eval=requireNamespace("nnet", quietly = TRUE)}
mmm_cov <- build_mmm(
  net,
  k = 2,
  n_starts = 3,
  max_iter = 60,
  seed = 1,
  covariates = c("project")
)

mmm_cov
```

# A Sensible Workflow

1. Build a transition network to prepare and inspect the sequence data.
2. Try a small set of defensible distances and clustering methods.
3. Use `cluster_choice()` to compare the candidates.
4. Inspect the chosen solution with `cluster_diagnostics()`.
5. Build per-cluster networks with `build_network(clustering)` or
   `cluster_network()`.
6. Use MMM as a second view when transition dynamics, not raw sequence
   similarity, are the main target.

