---
title: "Sequence Visualization and Pattern Comparison"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
options(max.print = 80)
library(Nestimate)
```

# Purpose

This tutorial covers two related tasks:

1. Visualizing raw sequences with `sequence_plot()`.
2. Comparing short sequence patterns with `sequence_compare()`.

The first task is descriptive. The second task is comparative and can include
statistical testing.

# Sequence Plots

The `trajectories` data set contains engagement states for 138 learners across
15 time points.

```{r trajectories-data}
data(trajectories)
dim(trajectories)
head(trajectories[, 1:8])
table(as.vector(trajectories), useNA = "ifany")
```

## Heatmap

The default plot is a clustered heatmap. Rows are learners and columns are
time points.

```{r heatmap-default}
sequence_plot(trajectories)
```

The default sorting uses sequence distance. Other sorting rules change the
question the plot answers.

```{r heatmap-sorts}
sequence_plot(trajectories, sort = "frequency",
              main = "Rows sorted by state frequency")

sequence_plot(trajectories, sort = "start",
              main = "Rows sorted by first observed state")
```

Use `k` to show broad row groups from the dendrogram.

```{r heatmap-k}
sequence_plot(
  trajectories,
  k = 3,
  k_color = "white",
  k_line_width = 2,
  main = "Sequence heatmap with three row groups"
)
```

## Index Plot

The index plot removes the dendrogram and gives each sequence one row.

```{r index-plot}
sequence_plot(trajectories, type = "index")
```

If you pass a `net_clustering`, `sequence_plot()` facets by cluster.

```{r cluster-for-plots}
cl <- build_clusters(
  as.data.frame(trajectories),
  k = 3,
  dissimilarity = "hamming",
  method = "ward.D2"
)

sequence_plot(cl, type = "index")
```

## Distribution Plot

The distribution view aggregates rows and shows how the state mix changes over
time.

```{r distribution-area}
sequence_plot(trajectories, type = "distribution")
```

Use bars and count scale when the number of observed learners matters.

```{r distribution-bars}
sequence_plot(
  trajectories,
  type = "distribution",
  geom = "bar",
  scale = "count"
)
```

The same view can be faceted by cluster.

```{r distribution-clustered}
sequence_plot(cl, type = "distribution")
```

## Custom Colors

Use named colors so the mapping stays clear.

```{r custom-colors}
engagement_colors <- c(
  Active = "#2A9D8F",
  Average = "#E9C46A",
  Disengaged = "#E76F51"
)

sequence_plot(
  trajectories,
  k = 3,
  state_colors = engagement_colors,
  legend = "bottom",
  legend_title = "Engagement",
  main = "Student engagement trajectories"
)
```

# Pattern Comparison

Sequence plots show the shape of the data. `sequence_compare()` tests whether
short patterns are distributed differently across groups.

We compare early and late parts of each human-AI session. The split is within
session, so the comparison asks whether patterns change over time inside the
same sessions.

```{r human-split}
data(human_long)
dat <- human_long[order(human_long$session_id, human_long$order_in_session), ]

dat$position <- ave(dat$order_in_session, dat$session_id, FUN = seq_along)
dat$n_events <- ave(dat$order_in_session, dat$session_id, FUN = length)
dat$half <- ifelse(dat$position <= dat$n_events / 2, "early", "late")

table(dat$half)
```

Build a grouped network using the broad `cluster` action labels.

```{r grouped-pattern-net}
pattern_net <- build_network(
  dat,
  method = "relative",
  action = "cluster",
  actor = "session_id",
  time = "order_in_session",
  group = "half"
)

pattern_net
```

Now compare 3- to 5-step patterns.

```{r sequence-compare}
patterns <- sequence_compare(
  pattern_net,
  sub = 3:5,
  min_freq = 20,
  test = "chisq",
  adjust = "fdr"
)

patterns
head(patterns$patterns, 12)
```

## Reading Residuals

For a given pattern, a positive residual in a group means the pattern appears
more often than expected in that group. A negative residual means it appears
less often than expected.

The pyramid plot works for two groups.

```{r pyramid}
plot(patterns, style = "pyramid", show_residuals = TRUE)
```

The heatmap is better when there are many groups or many patterns.

```{r heatmap-patterns}
plot(patterns, style = "heatmap")
```

Sort by frequency when the most common patterns are the focus.

```{r pyramid-frequency}
plot(patterns, style = "pyramid", sort = "frequency",
     show_residuals = TRUE)
```

# Test Choice

Use `test = "chisq"` when the groups are not independent, such as early and
late halves from the same sessions. Use `test = "permutation"` when group
labels are exchangeable across independent sequences.

The plot is only a summary. Always inspect `patterns$patterns` before making a
substantive claim.

