---
title: "State Frequencies and Chi-square Mosaics"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
options(max.print = 80)
library(Nestimate)
```

# Purpose

`plot_state_frequencies()` and `mosaic_plot()` look similar at first glance
because both use rectangles. They answer different questions.

`plot_state_frequencies()` asks: what states make up the data?

`mosaic_plot()` asks: which count cells are higher or lower than expected
under independence?

# State Frequencies for One Network

```{r single-network}
data(trajectories)

net <- build_network(
  as.data.frame(trajectories),
  method = "relative",
  format = "wide"
)

freq <- plot_state_frequencies(net)
freq
```

The returned object contains the plot and the tidy table.

```{r freq-table}
as.data.frame(freq)
```

The table is often the clearest place to read exact counts and proportions.

# State Frequencies by Group

```{r group-network}
data(group_regulation_long)

course_net <- build_network(
  group_regulation_long,
  method = "relative",
  actor = "Actor",
  action = "Action",
  time = "Time",
  group = "Course"
)

plot_state_frequencies(course_net)
```

The treemap emphasizes composition. A large tile means a state accounts for a
large share of that group.

Use bars when exact comparison by length is easier.

```{r state-bars}
plot_state_frequencies(course_net, style = "bars")
```

Switch the metric when counts matter more than proportions.

```{r state-bars-counts}
plot_state_frequencies(
  course_net,
  style = "bars",
  metric = "freq",
  label = "freq"
)
```

# Labels and Legends

```{r labels}
plot_state_frequencies(course_net, label = "all", legend = "right")
```

For dense panels, remove labels and let the legend carry the names.

```{r no-labels}
plot_state_frequencies(course_net, label = "none", legend = "bottom")
```

Abbreviation helps when names collide.

```{r abbreviate}
plot_state_frequencies(
  course_net,
  label = "all",
  abbreviate = TRUE,
  legend = "right"
)
```

# MCML State Frequencies

The function also dispatches on `mcml`. Each within-cluster layer becomes a
panel.

```{r mcml-object}
clusters <- list(
  Cognition = c("discuss", "synthesis", "consensus"),
  Metacognition = c("plan", "monitor", "adapt"),
  Social = c("coregulate", "cohesion", "emotion")
)

mc <- build_mcml(
  group_regulation_long,
  clusters = clusters,
  method = "sum",
  type = "frequency",
  actor = "Actor",
  time = "Time",
  action = "Action"
)

plot_state_frequencies(mc)
state_distribution(mc)
```

# Chi-square Mosaics

`mosaic_plot()` requires integer-valued weights. Use `method = "frequency"`
for transition counts.

```{r frequency-net}
data(ai_long)

ai_counts <- build_network(
  ai_long,
  method = "frequency",
  action = "code",
  actor = "session_id",
  time = "order_in_session"
)

ai_counts
```

```{r mosaic-single}
mosaic_plot(ai_counts, seed = 1)
```

In this plot, area is tied to count structure and color is the standardized
residual:

- blue: more transitions than expected;
- red: fewer transitions than expected;
- white: close to expected.

Permutation residuals are the default.

```{r mosaic-permutation}
mosaic_plot(ai_counts, residuals = "permutation", n_perm = 1000, seed = 1)
```

Use asymptotic residuals when comparing with tools based on
`chisq.test()$stdres`.

```{r mosaic-asymptotic}
mosaic_plot(ai_counts, residuals = "asymptotic")
```

Clip the color range when an extreme residual compresses the rest of the
scale.

```{r mosaic-range}
mosaic_plot(ai_counts, range = c(-4, 4), seed = 1)
```

# Group Mosaics

For `netobject_group`, `mosaic_plot()` builds one joint group-by-state mosaic.

```{r mosaic-group}
ai_group_counts <- build_network(
  ai_long,
  method = "frequency",
  action = "code",
  actor = "session_id",
  time = "order_in_session",
  group = "cluster"
)

mosaic_plot(ai_group_counts, seed = 1)
```

This plot asks whether groups distribute differently across states. It is not
one transition matrix per group.

# MCML Mosaics

The macro view shows cluster-to-cluster count residuals.

```{r mosaic-mcml-macro}
mosaic_plot(mc, level = "macro", seed = 1)
```

The cluster view shows one within-cluster mosaic per cluster.

```{r mosaic-mcml-clusters}
mosaic_plot(mc, level = "clusters", seed = 1)
```

# Choosing the Plot

Use `plot_state_frequencies()` for composition. Use `mosaic_plot()` for
residuals from count tables. If the network is row-normalized, use
frequencies or counts before making a mosaic.

