---
title: "Multi-Cluster Multi-Level Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
options(max.print = 80)
library(Nestimate)
```

# Purpose

MCML groups nodes into higher-level clusters. The result has a macro layer and
within-cluster layers.

Use MCML when the node set has meaningful groups, such as cognitive,
metacognitive, and social regulation actions.

# Data and Clusters

```{r data}
data(group_regulation_long)
head(group_regulation_long)
table(group_regulation_long$Action)
```

Define the node groups as a named list.

```{r clusters}
clusters <- list(
  Cognition = c("discuss", "synthesis", "consensus"),
  Metacognition = c("plan", "monitor", "adapt"),
  Social = c("coregulate", "cohesion", "emotion")
)

clusters
```

# Build MCML From Event Data

Use `build_mcml()` when starting from raw sequences or long event logs.

```{r build-mcml}
mc <- build_mcml(
  group_regulation_long,
  clusters = clusters,
  method = "sum",
  type = "tna",
  actor = "Actor",
  time = "Time",
  action = "Action"
)

mc
```

The macro layer is the cluster-level transition matrix.

```{r macro}
round(mc$macro$weights, 3)
mc$macro$inits
```

The within-cluster layers are stored in `mc$clusters`.

```{r within}
names(mc$clusters)
round(mc$clusters$Cognition$weights, 3)
round(mc$clusters$Metacognition$weights, 3)
round(mc$clusters$Social$weights, 3)
```

# Summary

```{r summary}
summary(mc)
```

`within_total` summarizes internal non-diagonal weight. `between_out` and
`between_in` summarize macro flow leaving and entering each cluster.

# Counts Instead of Probabilities

Use `type = "frequency"` when the analysis needs transition counts.

```{r build-frequency}
mc_counts <- build_mcml(
  group_regulation_long,
  clusters = clusters,
  method = "sum",
  type = "frequency",
  actor = "Actor",
  time = "Time",
  action = "Action"
)

mc_counts
round(mc_counts$macro$weights, 0)
```

Counts are the correct input for chi-square mosaics.

```{r mcml-mosaic}
mosaic_plot(mc_counts, level = "macro", seed = 1)
```

# Matrix Aggregation With `cluster_summary()`

Use `cluster_summary()` when a network matrix already exists. This path is
aggregation-only. It does not reconstruct a Markov process from raw sequences.

```{r cluster-summary}
net_counts <- build_network(
  group_regulation_long,
  method = "frequency",
  action = "Action",
  actor = "Actor",
  time = "Time"
)

mc_from_matrix <- cluster_summary(
  net_counts,
  clusters = clusters,
  method = "sum"
)

mc_from_matrix
round(mc_from_matrix$macro$weights, 0)
```

For matrix input, `method` controls arithmetic aggregation. Use `sum` for
counts, `mean` for already-normalized matrices, and robust summaries such as
`median` only when they match the question.

# Convert Layers to `netobject`s

`as_tna()` returns a `netobject_group`. The first element is `macro`; the rest
are within-cluster networks.

```{r as-tna}
mc_nets <- as_tna(mc)
mc_nets
names(mc_nets)
```

Once converted, each layer works with normal Nestimate tools.

```{r centrality}
net_centrality(mc_nets$macro)
```

```{r bootstrap-macro}
set.seed(1)
boot_macro <- bootstrap_network(mc_nets$macro, iter = 50)
boot_macro
```

# State Composition Inside MCML

State-frequency plots help explain what each cluster contains.

```{r state-frequency}
plot_state_frequencies(mc_counts)
```

The macro and within-cluster views answer different questions. The macro layer
describes movement among clusters; the within layers describe movement among
states inside each cluster.

# Labels

Use `labels` when node codes need readable names. The labels affect
within-cluster node display while macro labels remain cluster names.

```{r labels}
label_map <- c(
  discuss = "Discussion",
  synthesis = "Synthesis",
  consensus = "Consensus",
  plan = "Planning",
  monitor = "Monitoring",
  adapt = "Adapting",
  coregulate = "Co-regulation",
  cohesion = "Cohesion",
  emotion = "Emotion"
)

mc_labeled <- build_mcml(
  group_regulation_long,
  clusters = clusters,
  method = "sum",
  type = "frequency",
  actor = "Actor",
  time = "Time",
  action = "Action",
  labels = label_map
)

state_distribution(mc_labeled)
```

# Interpretation Checklist

When reading MCML output, separate four claims:

1. Which clusters are large in the state vocabulary?
2. Which clusters receive or send the most macro flow?
3. Which macro transitions are stable enough to interpret?
4. Which within-cluster transitions support a finer-grained claim?

Do not read macro edges as if they were node-level edges. They are aggregated
cluster-level transitions.
