---
title: "Markov, Higher-Order, and Simplicial Analysis"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
options(max.print = 80)
library(Nestimate)
```

# Purpose

This tutorial moves beyond a first look at a transition network. It covers:

- passage times and Markov stability;
- Markov order and higher-order pathways;
- path anomalies;
- simplicial summaries of network shape.

# Passage Times and Markov Stability

The `trajectories` data set has three engagement states over time.

```{r trajectories}
data(trajectories)
df <- as.data.frame(trajectories)

net <- build_network(df, method = "relative")
round(net$weights, 3)
```

`passage_time()` computes expected time to reach one state from another.

```{r passage-time}
pt <- passage_time(net)
pt
plot(pt)
```

The diagonal is the expected return time. Off-diagonal cells are expected
first passage times.

`markov_stability()` summarizes each state.

```{r markov-stability}
ms <- markov_stability(net)
ms
plot(ms)
```

Important quantities include persistence, stationary probability, return time,
sojourn time, and accessibility from other states.

## A Subgroup Comparison

Filter learners who are active more than half the observed time.

```{r active-subgroup}
active_count <- rowSums(df == "Active", na.rm = TRUE)
mostly_active <- df[active_count > ncol(df) / 2, ]

net_active <- build_network(mostly_active, method = "relative")
round(net_active$weights, 3)

passage_time(net_active)
markov_stability(net_active)
```

This is a descriptive subgroup comparison. For formal edge-level comparison,
use permutation tests on grouped networks.

# Higher-Order Analysis

Now use the human-AI action sequences. A first-order network asks what follows
the current action. Higher-order analysis asks whether earlier context changes
the next action.

```{r human-net}
data(human_long)

set.seed(3)
sessions <- sample(unique(human_long$session_id), 160)
human_sub <- human_long[human_long$session_id %in% sessions, ]

human_net <- build_network(
  human_sub,
  method = "relative",
  action = "code",
  actor = "session_id",
  time = "timestamp"
)

human_net
```

## Markov Order With MOGen

`build_mogen()` fits models at increasing order and reports information
criteria.

```{r mogen}
mg <- build_mogen(human_net, max_order = 3)
summary(mg)
plot(mg)
```

Extract order-specific transitions:

```{r mogen-transitions}
head(mogen_transitions(mg, order = 2), 12)
```

If a higher order is preferred, the next state depends on more than the
current state. If order 1 is preferred, local higher-order patterns may still
exist.

## Higher-Order Network

`build_hon()` identifies contexts where adding history changes the outgoing
transition distribution.

```{r hon}
hon <- build_hon(human_net)
hon

head(hon$ho_edges[order(-hon$ho_edges$count), ], 12)
```

Focus on higher-order edges where the context-specific probability differs
meaningfully from the first-order pattern.

## Common Paths

`path_counts()` is descriptive. It lists observed paths by frequency.

```{r path-counts}
path_counts(human_net, k = 3, top = 15)
```

Common paths are not automatically surprising. They may be common because
their individual states are common.

## Path Anomalies

`build_hypa()` tests which paths occur more or less often than expected.

```{r hypa}
hypa <- build_hypa(human_net)
hypa

head(hypa$over, 10)
head(hypa$under, 10)
```

Over-represented paths are routines that occur more than the null model
expects. Under-represented paths are sequences that are avoided relative to
that expectation.

# Simplicial Analysis

Simplicial analysis treats groups of connected nodes as higher-dimensional
objects. It is a way to describe the shape of the network, not just the list
of edges.

```{r simplicial}
sc <- build_simplicial(human_net, threshold = 0.05)
sc

betti_numbers(sc)
euler_characteristic(sc)
simplicial_degree(sc)
```

Read Betti numbers as:

- beta 0: connected components;
- beta 1: loops;
- beta 2: enclosed voids.

```{r plot-simplicial}
plot(sc)
```

Persistent homology varies the threshold and tracks which topological features
persist.

```{r persistent}
ph <- persistent_homology(human_net, n_steps = 20)
ph
plot(ph)
```

Q-analysis studies connectivity through shared faces.

```{r q-analysis}
qa <- q_analysis(sc)
qa
plot(qa)
```

# Practical Reading Order

Use the methods in this order:

1. Estimate and validate a first-order transition network.
2. Use `passage_time()` and `markov_stability()` to describe movement.
3. Use `build_mogen()` to check whether higher order is needed globally.
4. Use `build_hon()` and `path_counts()` to inspect local pathways.
5. Use `build_hypa()` when the question is statistical surprise.
6. Use simplicial tools when the question is network shape.

